### 2022-12-11 paper/paper.tex
added stack template and paper setup

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
%
\title{Alternative Methods for Implementing
    Explicit and Finding Implicit Sharing in embedded DSLs}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Curtis D'Alves \and
Christopher Anand \and
Lucas Dutton \and
Steven Gonder
}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{McMaster University, 1280 Main St W Hamilton, Canada}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
TODO The abstract should briefly summarize the contents of the paper in
150--250 words.

\keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\section{Introduction}

\section{Examples TODO REMOVE ME}
example citiation \cite{kiselyov:sharing}
\begin{table}
\caption{Table captions should be placed above the
tables.}\label{tab1}
\begin{tabular}{|l|l|l|}
\hline
Heading level &  Example & Font size and style\\
\hline
Title (centered) &  {\Large\bfseries Lecture Notes} & 14 point, bold\\
1st-level heading &  {\large\bfseries 1 Introduction} & 12 point, bold\\
2nd-level heading & {\bfseries 2.1 Printing Area} & 10 point, bold\\
3rd-level heading & {\bfseries Run-in Heading in Bold.} Text follows & 10 point, bold\\
4th-level heading & {\itshape Lowest Level Heading.} Text follows & 10 point, italic\\
\hline
\end{tabular}
\end{table}

\noindent Displayed equations are centered and set on a separate
line.
\begin{equation}
x + y = z
\end{equation}
Please try to avoid rasterized images for line-art diagrams and
schemas. Whenever possible, use vector graphics instead (see
Fig.~\ref{fig1}).

\begin{figure}
\includegraphics[width=\textwidth]{fig1.eps}
\caption{A figure caption is always placed below the illustration.
Please note that short captions are centered, while long ones are
justified by the macro package automatically.} \label{fig1}
\end{figure}

\subsubsection{Acknowledgements} Please place your acknowledgments at
the end of the paper, preceded by an unnumbered run-in heading (i.e.
3rd-level heading).

\bibliographystyle{splncs04}
\bibliography{references}
%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
% \bibliographystyle{splncs04}
% \bibliography{mybibliography}
%
% \begin{thebibliography}{8}
% \bibitem{ref_article1}
% Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)

% \bibitem{ref_lncs1}
% Author, F., Author, S.: Title of a proceedings paper. In: Editor,
% F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
% Springer, Heidelberg (2016). \doi{10.10007/1234567890}

% \bibitem{ref_book1}
% Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
% Location (1999)

% \bibitem{ref_proc1}
% Author, A.-B.: Contribution title. In: 9th International Proceedings
% on Proceedings, pp. 1--2. Publisher, Location (2010)

% \bibitem{ref_url1}
% LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
% Oct 2017
% \end{thebibliography}
\end{document}

### 2022-12-12 paper/paper.tex
initial paper/prototype setup

@@
 
-\section{Examples TODO REMOVE ME}
-example citiation \cite{kiselyov:sharing}
-\begin{table}
-\caption{Table captions should be placed above the
-tables.}\label{tab1}
-\begin{tabular}{|l|l|l|}
-\hline
-Heading level &  Example & Font size and style\\
-\hline
-Title (centered) &  {\Large\bfseries Lecture Notes} & 14 point, bold\\
-1st-level heading &  {\large\bfseries 1 Introduction} & 12 point, bold\\
-2nd-level heading & {\bfseries 2.1 Printing Area} & 10 point, bold\\
-3rd-level heading & {\bfseries Run-in Heading in Bold.} Text follows & 10 point, bold\\
-4th-level heading & {\itshape Lowest Level Heading.} Text follows & 10 point, italic\\
-\hline
-\end{tabular}
-\end{table}
+TODO describe sharing problem (mention observable sharing
+\cite{gill:observablesharing} and implicit/explicit sharing
+\cite{kiselyov:sharing} papers)
 
-\noindent Displayed equations are centered and set on a separate
-line.
-\begin{equation}
-x + y = z
-\end{equation}
-Please try to avoid rasterized images for line-art diagrams and
-schemas. Whenever possible, use vector graphics instead (see
-Fig.~\ref{fig1}).
+TODO describe finally tagless \cite{carette:finallytagless}
 
-\begin{figure}
-\includegraphics[width=\textwidth]{fig1.eps}
-\caption{A figure caption is always placed below the illustration.
-Please note that short captions are centered, while long ones are
-justified by the macro package automatically.} \label{fig1}
-\end{figure}
+TODO overview of our method
+
+\section{Detecting Sharing In Finally Tagless DSLs}
+TODO
+
+\section{Implicit Sharing Via Hash-Consing}
+TODO
+cite Ershov’s original description of hash-consing \cite{ershov1958:consing}
+cite Type safe consing implementation (with performance benchamrks) \cite{filliatre:typesafeconsing}
+
+\section{Limitations of Hash-Consing}
+
+TODO add-chains example
+
+\subsection{Explicit Sharing and Limitations}
+TODO
+
+\section{Implicit Sharing Via ByteString ASTs}
+TODO
+
+\subsection{Memory Limitations}
+TODO
+\section{Explicit Sharing Of ByteString ASTs}
+TODO
 

### 2022-12-12 AltSharing/app/Main.hs

@@ -1,6 +1,6 @@
-module Main (main) where
+module Main where
 
-import Lib
+import Lib ()
 
 main :: IO ()
-main = someFunc
+main = putStrLn "Hello World"

### 2022-12-12 AltSharing/src/Lib.hs

@@ -1,6 +1,34 @@
-module Lib
-    ( someFunc
-    ) where
+module Lib where
 
-someFunc :: IO ()
-someFunc = putStrLn "someFunc"
+import Data.Word
+import Control.Monad.State.Strict
+
+import Data.Graph.Inductive.Graph
+import Data.Graph.Inductive.PatriciaTree (Gr)
+
+-- * Core DSL
+
+data GPR = GPR { unGPR :: Word64 }
+data VR = VR { unVR :: (Word64,Word64) }
+
+class CoreISA repr where
+  toGPR :: Word64 -> repr Word64
+  add :: repr Word64 -> repr Word64 -> repr Word64
+
+-- * Interpretation
+
+newtype Interp a = Interp { runInterp :: a }
+  deriving (Show,Eq)
+
+instance CoreISA Interp where
+  toGPR w = Interp w
+  add g0 g1 = Interp $ runInterp g0 + runInterp g1
+
+-- * Graph Generation
+
+data OpLabel = GPROp Word64
+             | ADDOp
+
+type NodeID = Int
+
+type Graph = Gr OpLabel ()

### 2022-12-13 paper/paper.tex
still working on initial prototype / description of sharing problem

@@
 \usepackage{graphicx}
+\usepackage{listings}
 % Used for displaying a sample figure. If possible, figure files should
@@
 \begin{document}
+\lstset{language=haskell}
 %
@@
 
-TODO overview of our method
+We present methods for implementing embedded DSLs with sharing that are both
+safe and maintain all the benefits of being embedded in the Haskell ecosystem.
+This means DSL functions are type-safe, do not require the use of unsafe
+referencing (i.e., via unsafePerformIO) and can return Haskell's container
+types (i.e., tuples, lists, etc) without breaking sharing.
 
-\section{Detecting Sharing In Finally Tagless DSLs}
-TODO
+\section{Detecting Sharing}
 
-\section{Implicit Sharing Via Hash-Consing}
+A naive DSL implementation of an expression in Haskell can be done via standard
+Haskell data types, for example:
+
+\begin{lstlisting}
+data Exp
+  = Add Exp Exp
+  | Variable String
+  | Constant Int
+
+-- Example
+v0 = Variable ``v0''
+exp0 = Add v0 (Constant 0)
+exp1 = Add exp0 exp0
+\end{lstlisting}
+
+Note the DSL generates a tree, or to be more specific an Abstract Syntax Tree
+(AST).
+
+\subsection{Detecting Sharing In Finally Tagless DSLs}
+
+\subsection{Implicit Sharing Via Hash-Consing}
 TODO
 cite Ershov’s original description of hash-consing \cite{ershov1958:consing}
-cite Type safe consing implementation (with performance benchamrks) \cite{filliatre:typesafeconsing}
+cite Type safe consing implementation (with performance benchmarks) \cite{filliatre:typesafeconsing}
 
-\section{Limitations of Hash-Consing}
+\subsection{Limitations of Hash-Consing}
 
@@
 
-\bibliographystyle{splncs04}
-\bibliography{references}
 %
@@
 % References will then be sorted and formatted in the correct style.
-%
-% \bibliographystyle{splncs04}
-% \bibliography{mybibliography}
-%
-% \begin{thebibliography}{8}
-% \bibitem{ref_article1}
-% Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
+\bibliographystyle{splncs04}
+\bibliography{references}
 
-% \bibitem{ref_lncs1}
-% Author, F., Author, S.: Title of a proceedings paper. In: Editor,
-% F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
-% Springer, Heidelberg (2016). \doi{10.10007/1234567890}
+\end{document}
 
-% \bibitem{ref_book1}
-% Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
-% Location (1999)
-
-% \bibitem{ref_proc1}
-% Author, A.-B.: Contribution title. In: 9th International Proceedings
-% on Proceedings, pp. 1--2. Publisher, Location (2010)
-
-% \bibitem{ref_url1}
-% LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
-% Oct 2017
-% \end{thebibliography}
-\end{document}
+% LocalWords:  DSLs unsafePerformIO Haskell's

### 2022-12-13 AltSharing/src/Lib.hs

@@ -3,8 +3,11 @@ module Lib where
 import Data.Word
 import Control.Monad.State.Strict
 
-import Data.Graph.Inductive.Graph
+import Data.Graph.Inductive.Graph ()
 import Data.Graph.Inductive.PatriciaTree (Gr)
+
+import Data.HashMap.Strict (HashMap)
+import qualified Data.HashMap.Strict as HashMap
 
 -- * Core DSL
 
@@ -21,7 +24,7 @@ newtype Interp a = Interp { runInterp :: a }
   deriving (Show,Eq)
 
 instance CoreISA Interp where
-  toGPR w = Interp w
+  toGPR = Interp
   add g0 g1 = Interp $ runInterp g0 + runInterp g1
 
 -- * Graph Generation
@@ -31,4 +34,9 @@ data OpLabel = GPROp Word64
 
 type NodeID = Int
 
-type Graph = Gr OpLabel ()
+data GData = GData { gGraph :: Gr OpLabel ()
+                   , gMap :: HashMap (OpLabel,[NodeID]) NodeID
+                   , gCounter :: Int
+                   }
+
+newtype Graph a = Graph { unGraph :: State GData NodeID }

### 2022-12-13 paper/paper.tex
working on background

@@
 \usepackage{listings}
+\usepackage{minted}
 % Used for displaying a sample figure. If possible, figure files should
@@
 
-\begin{lstlisting}
+\begin{minted}{haskell}
 data Exp
@@
 exp1 = Add exp0 exp0
-\end{lstlisting}
+\end{minted}
 
 Note the DSL generates a tree, or to be more specific an Abstract Syntax Tree
-(AST).
+(AST). Common features a DSL implementer might implement would include code
+generation or pretty printing. Simple traversal of the AST for either of these
+operations would result in duplication, for example in the above code snippet
+the AST for {\bf exp0} will be traversed twice. For code generation in
+particular this would be problematic, in order to circumvent this problem in
+general we need to perform common subexpression elimination by converting the
+AST into a directed acyclic graph. % #TODO cite literature
+
+TODO describe DAG conversion by pointer comparison in the state monad
 
 \subsection{Detecting Sharing In Finally Tagless DSLs}
+
+Monads are useful, but don't make for a very user friendly DSL. It would be nice
+to make use of monadic state when we need it (i.e., for converting to a DAG)
+while hiding it behind a nice pure interface. The final tagless approach of
+\cite{carette:finallytagless} is popular for accomplishing this. In this
+approach, DSL expressions are built using typeclass methods that wrap the DSL in
+a parameterized representation. For example, the previous data type based DSL
+could be written in finally tagless as
+
+\begin{minted}{haskell}
+class Exp repr where
+  add :: repr Int -> repr Int -> repr Int
+  variable :: String -> repr Int
+  constant :: Int -> repr Int
+\end{minted}
+
+We can then create different instances to implement different functionality.
+For example, we can generate the AST from the previous DSL like so
+\begin{minted}{haskell}
+data ExpR a = ExpR { unExpR :: Exp }
+
+instance Exp ExpR where
+  constant = ExpR . Constant
+  variable = ExpR . Variable
+  add (ExpR x) (ExpR y) = ExpR (Add x y)
+\end{minted}
+
+Or we can implement pretty printing
+\begin{minted}{haskell}
+newtype Pretty a = Pretty { unPretty :: String }
+
+instance Exp Pretty where
+  add x y = Pretty $ "("++unPretty x++") + ("++unPretty y++")"
+  variable x = Pretty x
+  constant x = Pretty $ show x
+\end{minted}
+
+And use the same DSL code to run either implementation
+\begin{minted}{haskell}
+exp :: repr Int -> repr Int
+exp v0 =
+  let
+    exp0 = add v0 (constant 0)
+  in add exp0 exp0
+
+expR = unExpR $ exp $ variable "v0"
+expP = unPretty $ exp $ variable "v0"
+\end{minted}
+
+Finally tagless style provides extensible, user friendly DSLs. However there
+are still some complications when using it to implement sharing.
 
@@
 
+In \cite{kiselyov:sharing}, a solution for detection implicit sharing in finally
+tagless style is presented via the method of hash-consing. You can find a more
+throrough explanation of the method there, but we'll give an overview here. This
+method first involves defining a DAG type, for example
+\begin{minted}{haskell}
+type NodeID = Int
+data Node = NAdd NodeID NodeID
+          | NVariable String
+          | NConstant Int
+
+data BiMap a -- abstract
+lookup_key : Ord a => a -> BiMap a -> Maybe Int
+lookup_val : Int -> BiMap a -> a
+insert : Ord a => a -> BiMap a -> (Int,BiMap a)
+empty : BiMap a
+
+newtype DAG = DAG (BiMap Node) deriving Show
+\end{minted}
+
+Note the purpose of the BiMap type is to be able to quickly insert and lookup
+nodes by their NodeID (i.e., a bijection of Node's and their NodeID's), and is
+most optimally implemented as a hash table with linear probing. The
+representation for the finally tagless instance is then a wrapper around a State
+monad that holds DAG in its state and returns the current (top) NodeID.
+
+\begin{minted}{haskell}
+newtype Graph a = Graph { unGraph :: State DAG NodeID }
+
+instance Exp Graph where
+  constant x = Graph (hashcons $ NConstant x)
+  variable x = Graph (hashcons $ NVariable x)
+  add e1 e2 = Graph (do
+                     h1 <- unGraph e1
+                     h2 <- unGraph e2
+                     hashcons $ NAdd h1 h2)
+\end{minted}
+
+The trick to uncovering sharing in the implementation is implemented via the
+{\bf hashcons} function, which inserts a new node into the current DAG, but not
+before checking if it is already there.
+\begin{minted}{haskell}
+hashcons :: Node -> State DAG NodeID
+hashcons e = do
+  DAG m <- get
+  case lookup_key e m of
+    Nothing -> let (k,m') = insert e m
+               in put (DAG m') >> return k
+    Just k -> return k
+\end{minted}
+
 \subsection{Limitations of Hash-Consing}
 
-TODO add-chains example
+When we wrap our State monad in finally tagless style, we lose some of Haskell's
+built-in sharing capability. Consider the following code, note that the use of
+local variables explicitely defines the computation $x + y$ to only be computed
+once
+\begin{minted}{haskell}
+haskellSharing x y=
+ let
+   z = x + y
+ in z + z
+\end{minted}
+
+Implicit sharing via hash-consing prevents duplication in the resulting DAG, but
+unfortunately doesn't prevent redundant computation. Consider the following
+equivalent attempt at using Haskell's built-in sharing in the finally tagless DSL
+\begin{minted}{haskell}
+redundant :: Exp Graph -> Exp Graph -> Exp Graph
+redundant x y =
+  let
+    z = add x y
+  in add z z
+\end{minted}
+Note {\bf z} is a wrapper around a State monad. Recall the implementation of
+{\bf add} via hash consing
+\begin{minted}{haskell}
+add e1 e2 = Graph (do
+                   h1 <- unGraph e1
+                   h2 <- unGraph e2
+                   hashcons $ NAdd h1 h2)
+\end{minted}
+In order to
+
+TODO describe add-chains example
+
+TODO translate to kiselyov's language
+\begin{minted}{haskell}
+addChains :: forall repr. CoreISA repr => (repr VR, repr VR, repr VR, repr VR) -> (repr VR, repr VR, repr VR, repr VR)
+addChains =
+      add2 . add1
+    . add2 . add1
+    . add2 . add1
+    . add2 . add1
+    . add2 . add1
+\end{minted}
 
@@
 
-% LocalWords:  DSLs unsafePerformIO Haskell's
+% LocalWords:  DSLs unsafePerformIO Haskell's AST typeclass tagless
+% Local Variables:
+% LaTeX-verbatim-environments-local: ("minted")
+% eval: (setq-local LaTeX-indent-environment-list (cons '("minted" current-indentation) (default-value 'LaTeX-indent-environment-list)))
+% End:

### 2022-12-13 AltSharing/src/Lib.hs

@@ -40,3 +40,7 @@ data GData = GData { gGraph :: Gr OpLabel ()
                    }
 
 newtype Graph a = Graph { unGraph :: State GData NodeID }
+
+
+-- instance CoreISA Graph where
+--   toGPR x = 

### 2022-12-13 paper/paper.tex
working on issues with hash-consing in prototype and paper

@@
 \begin{minted}{haskell}
-data ExpR a = ExpR { unExpR :: Exp }
+newtype ExpR a = ExpR { unExpR :: Exp }
 
@@
 \begin{minted}{haskell}
-exp :: repr Int -> repr Int
+exp :: Exp repr => repr Int -> repr Int
 exp v0 =
@@
 data BiMap a -- abstract
-lookup_key : Ord a => a -> BiMap a -> Maybe Int
-lookup_val : Int -> BiMap a -> a
-insert : Ord a => a -> BiMap a -> (Int,BiMap a)
-empty : BiMap a
+lookup_key :: Ord a => a -> BiMap a -> Maybe Int
+lookup_val :: Int -> BiMap a -> a
+insert :: Ord a => a -> BiMap a -> (Int,BiMap a)
+empty :: BiMap a
 

### 2022-12-13 AltSharing/src/HashConsSharing.hs

@@ -1,0 +1,119 @@
+{- This code is based of the work of Oleg, Kiselyov's
+ - "Implementing explicit and finding implicit sharing in embedded DSLs"
+ -}
+module HashConsSharing where
+
+import Prelude hiding (exp)
+import qualified Data.Map.Strict as Map
+import Data.Map.Strict (Map)
+import Debug.Trace
+import Data.Word
+import Control.Monad.State.Strict
+
+-- * Data Type based DSL
+
+-- | Data Type based Expression DSL
+data Exp0
+  = Add Exp0 Exp0
+  | Variable String
+  | Constant Int
+
+-- Example
+v0 = Variable "v0"
+exp0 = Add v0 (Constant 0)
+exp1 = Add exp0 exp0
+
+-- * Finally Tagless Style DSL
+
+-- | Finally Tagless Style Expression DSL
+class Exp repr where
+  add :: repr Int -> repr Int -> repr Int
+  variable :: String -> repr Int
+  constant :: Int -> repr Int
+
+-- | AST Representation (no-sharing)
+newtype ExpR a = ExpR { unExpR :: Exp0 }
+
+instance Exp ExpR where
+  constant = ExpR . Constant
+  variable = ExpR . Variable
+  add (ExpR x) (ExpR y) = ExpR (Add x y)
+
+
+-- | Pretty Printing representation
+newtype Pretty a = Pretty { unPretty :: String }
+
+instance Exp Pretty where
+  add x y = Pretty $ "("++unPretty x++") + ("++unPretty y++")"
+  variable x = Pretty x
+  constant x = Pretty $ show x
+
+-- | Example usage of finally tagless
+exp :: Exp repr => repr Int -> repr Int
+exp v0 =
+  let
+    exp0 = add v0 (constant 0)
+  in add exp0 exp0
+
+expR = unExpR $ exp $ variable "v0"
+expP = unPretty $ exp $ variable "v0"
+
+-- * Hash Consing
+
+-- | A node is identified by a unique Int
+type NodeID = Int
+
+-- | Each Node has a constructor that holds either literals or the NodeID of its
+-- inputs, making it unique so long as NodeID's are unique
+data Node = NAdd NodeID NodeID
+          | NVariable String
+          | NConstant Int
+ deriving (Show,Eq,Ord)
+
+
+-- | Implementation of BiMap as a strict Map that also keeps track of the max nodeID
+data BiMap a = BiMap { biMap :: Map a NodeID
+                     , maxNodeID :: NodeID
[…]

### 2022-12-14 paper/paper.tex
plotting addchains hashconsing scaling example

@@
 \begin{minted}{haskell}
-redundant :: Exp Graph -> Exp Graph -> Exp Graph
-redundant x y =
+dslSharing :: Exp Graph -> Exp Graph -> Exp Graph
+dslSharing x y =
   let
@@
 \end{minted}
-In order to
+The values {\bf h1} and {\bf h2} need to be explicitely evaluated through the
+State monad, meaning even if {\bf e1} and {\bf e2} are the same shared Haskell
+value, their underlying computations will be performed twice. Hash-consing will
+prevent these redundencies from appearing in the resulting DAG, the the entire
+unshared AST wil still be traversed, performing a hash-cons on each node.
 

### 2022-12-14 AltSharing/src/HashConsSharing.hs

@@ -6,9 +6,9 @@ module HashConsSharing where
 import Prelude hiding (exp)
 import qualified Data.Map.Strict as Map
 import Data.Map.Strict (Map)
-import Debug.Trace
 import Data.Word
 import Control.Monad.State.Strict
+import Graphics.EasyPlot
 
 -- * Data Type based DSL
 
@@ -23,6 +23,7 @@ v0 = Variable "v0"
 exp0 = Add v0 (Constant 0)
 exp1 = Add exp0 exp0
 
+-- 
 -- * Finally Tagless Style DSL
 
 -- | Finally Tagless Style Expression DSL
@@ -94,7 +95,9 @@ empty :: BiMap a
 empty = BiMap Map.empty 0
 
 -- | A Directed Acyclic Graph is a BiMap between Node and NodeID
-newtype DAG = DAG (BiMap Node) deriving Show
+data DAG = DAG { dagMAP :: BiMap Node  -- | A map from Node to NodeID
+               , dagCnt :: Int -- | used to track # of hashcons performed
+               } deriving Show
 
 -- | DAG construction representation via the State monad
 newtype Graph a = Graph { unGraph :: State DAG NodeID }
@@ -107,13 +110,32 @@ instance Exp Graph where
                      h2 <- unGraph e2
                      hashcons $ NAdd h1 h2)
 
+buildDAG g = runState (unGraph g) (DAG empty 0)
+
 -- | naive implementation of hash consing, insert a node into the DAG only if it isn't
 -- already there
 hashcons :: Node -> State DAG NodeID
 hashcons e = do
-  DAG m <- get
-  traceM $ "hashcons performed"
+  DAG m cnt <- get
+  modify (\dag -> dag { dagCnt = cnt+1}) -- track number of hashcons performed
   case lookup_key e m of
     Nothing -> let (k,m') = insert e m
-               in put (DAG m') >> return k
+               in modify (\dag -> dag { dagMAP = m' })
+                  >> return k
     Just k -> return k
+
+
+-- | Example of exponential scale of hash-consing
+addChains :: Exp repr => Int -> repr Int -> repr Int
+addChains n x0 = head $ drop n $ iterate (\x -> add x x) x0
+
+plotAddChains size =
+  let
+    chainsData = map (\(x,y) -> (fromIntegral x,fromIntegral y))
[…]

### 2022-12-14 paper/paper.tex
background done

@@
 value, their underlying computations will be performed twice. Hash-consing will
-prevent these redundencies from appearing in the resulting DAG, the the entire
+prevent these redundencies from appearing in the resulting DAG, but the entire
 unshared AST wil still be traversed, performing a hash-cons on each node.
 
-TODO describe add-chains example
-
-TODO translate to kiselyov's language
+Consider a chain of {\bf add}'s  with sharing, for example
 \begin{minted}{haskell}
-addChains :: forall repr. CoreISA repr => (repr VR, repr VR, repr VR, repr VR) -> (repr VR, repr VR, repr VR, repr VR)
-addChains =
-      add2 . add1
-    . add2 . add1
-    . add2 . add1
-    . add2 . add1
-    . add2 . add1
+addChains :: Exp repr => Expr Int -> Expr Int
+addChains x0 = let
+    x1 = add x0 x0
+    x2 = add x1 x1
+    ...
+  in xn
 \end{minted}
 
+\begin{figure}
+  \centering
+  \includegraphics[width=0.6\textwidth]{figs/hashcons.png}
+  \caption{Number of hashcons to add operations performed} \label{fig:hashcons}
+\end{figure}
+As you can see from figure \ref{fig:hashcons}, this code will perform approximately
+$2^{n+1}$ hashcons operations, where n is the number of {\bf add}'s.
+
 \subsection{Explicit Sharing and Limitations}
-TODO
+
+\cite{kiselyov:sharing} acknowledges the amount of computation with hash-consing
+can get out of control, and proposes an ad-hoc solution, explicit sharing via a
+custom let construct
+\begin{minted}{haskell}
+class ExpLet repr where
+  let_ :: repr a -> (repr a -> repr b) -> repr b
+instance ExpLet Graph where
+  let_ e f = Graph (do x <- unGraph e
+                     unGraph $ f (Graph (return x)))
+
+addChains x =
+  let_ x (\x0 ->
+  let_ (add x0 x0)  (\x1 ->
+  let_ (add x1 x1)  (\x2 ->
+   ...
+  )))
+\end{minted}
+This makes the code a bit clunky and adds an extra burden on the DSL
+writer, but it prevents unnecessary hash-consing in our example. However the
+method has it's limitations, suppose we want to extend our DSL to support
+functions that return multiple outputs, such as tuples or container types like
+lists (for example \mintinline{haskell}{vadd :: (repr Int,repr Int) -> (repr Int,repr
+  Int)}). First of all, we need to implement different versions of the custom let
+construct to correspond to the number of outputs
+\begin{minted}{haskell}
+class ExpLet repr where
+  let_ :: repr a -> (repr a -> repr b) -> repr b
+  let_ 2 :: repr a -> (repr a -> (repr b,repr c)) -> (repr b,repr c)
+
+instance ExpLet repr where
+   ...
+   let_2 e f = let
+       g0 = Graph (do x <- unGraph e
+                        let (o0,o1) = f (Graph (return x))
+                        unGraph o0)
+       g1 = Graph (do x <- unGraph e
+                        let (o0,o1) = f (Graph (return x))
+                        unGraph o1)
+       in (g0,g1)
+\end{minted}
+If need be, its possible to enumerate custom let instances for every amount of
+outputs or container types we would need, we could even use template Haskell to
+accomplish this. However, this custom let construct now has a new source of
+redundancy in its outputs. Each output it returns will now have to individually
+evaluate it's input, so a chain of DSL functions that output 2 or more values
+will suffer from the same exponential scaling of hashcons.
+
+One solution to this issue is to integrate container types such as tuples and lists
+into the DSL language. However doing this will essentially make the DSL ``less
+embedded'', manipulating tuple values will be cumbersome constantly requiring
+calls to custom implementations of {\bf fst}/{\bf snd} etc. And for lists you'll
+lose all access to built-in Haskell list functionality.
 

### 2022-12-14 AltSharing/src/HashConsSharing.hs

@@ -133,9 +133,18 @@ plotAddChains size =
   let
     chainsData = map (\(x,y) -> (fromIntegral x,fromIntegral y))
       [ (n,dagCnt $ snd $ buildDAG $ addChains n $ variable "x") | n <- [0..size] ]
-  in plot (PNG "plot.png") $ Data2D [Title "Hashcons Scaling",Style Lines] [] chainsData
+  in plot (PNG "plot.png") $ Data2D [Title "Hashcons Scaling",Style Linespoints,Color Red] [] chainsData
 
-  -- let
-  --   x1 = add x0 x0
-  --   x2 = add x1 x1
-  -- in x2
+
+-- * Explicit Sharing
+
+class ExpLet repr where
+  let_ :: repr a -> (repr a -> repr b) -> repr b
+
+instance ExpLet Graph where
+  let_ e f = Graph (do x <- unGraph e
+                       unGraph $ f (Graph (return x)))
+
+-- | Example of explicit sharing
+addChainsE :: (Exp repr,ExpLet repr) => Int -> repr Int -> repr Int
+addChainsE n x0 = head $ drop n $ iterate (\xn -> let_ xn (\x -> add x x)) x0

### 2022-12-15 paper/paper.tex
added prototypes of bytestring sharing and bytestring caching

@@
 %
-%\titlerunning{Abbreviated paper title}
+\titlerunning{Alternative Explicit and Implicit Sharing}
 % If the paper title is too long for the running head, you can set
@@
 %
-\authorrunning{F. Author et al.}
+\authorrunning{C. D'Alves et al.}
 % First names are abbreviated in the running head.
@@
 \end{minted}
-This makes the code a bit clunky and adds an extra burden on the DSL
-writer, but it prevents unnecessary hash-consing in our example. However the
-method has it's limitations, suppose we want to extend our DSL to support
-functions that return multiple outputs, such as tuples or container types like
-lists (for example \mintinline{haskell}{vadd :: (repr Int,repr Int) -> (repr Int,repr
-  Int)}). First of all, we need to implement different versions of the custom let
-construct to correspond to the number of outputs
+This makes the code a bit clunky and adds an extra burden on the DSL writer, but
+it prevents unnecessary hash-consing in our example. However the method has it's
+limitations, suppose we want to write a DSL function that returns multiple
+outputs, such as tuples or container types like lists (for example
+\mintinline{haskell}{vadd :: (repr Int,repr Int) -> (repr Int,repr Int)}). %TODO
+                                % need to give implementation of this function?
+First of all, we need to implement different versions of the custom let construct to
+correspond to the number of outputs
 \begin{minted}{haskell}
@@
 
-One solution to this issue is to integrate container types such as tuples and lists
-into the DSL language. However doing this will essentially make the DSL ``less
-embedded'', manipulating tuple values will be cumbersome constantly requiring
-calls to custom implementations of {\bf fst}/{\bf snd} etc. And for lists you'll
-lose all access to built-in Haskell list functionality.
+One solution to this issue is to integrate container types such as tuples and
+lists into the DSL language. However doing this will take away form the
+advantages of having an embedded language, manipulating tuple values will be
+cumbersome constantly requiring calls to custom implementations of {\bf
+  fst}/{\bf snd} etc. And for lists you'll lose all access to built-in Haskell
+list functionality.
 

### 2022-12-15 AltSharing/src/ByteStringExplicit.hs

@@ -1,0 +1,114 @@
+{-# LANGUAGE OverloadedStrings #-}
+
+module ByteStringExplicit where
+
+import Prelude hiding (exp)
+import qualified Data.Map.Strict as Map
+import Data.Map.Strict (Map)
+import Control.Monad.State.Strict
+import Graphics.EasyPlot
+import Data.ByteString.Char8 (ByteString)
+import qualified Data.ByteString.Char8 as ByteString
+import Data.Trie (Trie)
+import qualified Data.Trie as Trie
+import qualified Data.List as List
+
+-- | Finally Tagless Style Expression DSL
+class Exp repr where
+  add :: repr Int -> repr Int -> repr Int
+  variable :: String -> repr Int
+  constant :: Int -> repr Int
+
+-- | Explicit caching of ByteStrings
+class Cacheable repr where
+  cache :: ByteString -> repr ByteString -> repr ByteString
+
+instance Cacheable Graph where
+  cache s' (Graph g s _) = Graph g s' (Just s)
+
+-- | A node is identified by a unique Int
+type NodeID = Int
+
+-- | Each Node has a constructor that holds either literals or the NodeID of its
+-- inputs, making it unique so long as NodeID's are unique
+data Node = NAdd NodeID NodeID
+          | NVariable String
+          | NConstant Int
+ deriving (Show,Eq,Ord)
+
+
+-- | A Directed Acyclic Graph is inside the values of the Trie
+data DAG = DAG { dagTrie :: Trie (Node,NodeID) -- | a trie from bytestrings to Node/NodeID
+               , dagCache :: Trie ByteString -- | FIXME switch to Map?
+               , dagMaxID :: Int -- | used to track # of hashcons performed
+               } deriving Show
+
+-- | DAG construction representation via the State monad
+data Graph a = Graph { unGraph :: State DAG NodeID
+                     , unStringAST :: ByteString
+                     , unAddCache :: Maybe ByteString }
+
+-- | Build a AST encoded as a ByteString
+buildStringAST :: Node -> [ByteString] -> ByteString
+buildStringAST node args =
+  let
+    opString = case node of
+                 NAdd _ _ -> "nadd"
+                 NVariable s -> "$" <> ByteString.pack s
+                 NConstant c -> "#" <> (ByteString.pack $ show c)
+    argsString = "(" <> ByteString.concat (List.intersperse "," args) <> ")"
[…]

### 2022-12-15 AltSharing/src/ByteStringSharing.hs

@@ -1,0 +1,94 @@
+{-# LANGUAGE OverloadedStrings #-}
+
+module ByteStringSharing where
+
+import Prelude hiding (exp)
+import qualified Data.Map.Strict as Map
+import Data.Map.Strict (Map)
+import Control.Monad.State.Strict
+import Graphics.EasyPlot
+import Data.ByteString.Char8 (ByteString)
+import qualified Data.ByteString.Char8 as ByteString
+import Data.Trie (Trie)
+import qualified Data.Trie as Trie
+import qualified Data.List as List
+
+-- | Finally Tagless Style Expression DSL
+class Exp repr where
+  add :: repr Int -> repr Int -> repr Int
+  variable :: String -> repr Int
+  constant :: Int -> repr Int
+
+-- | A node is identified by a unique Int
+type NodeID = Int
+
+-- | Each Node has a constructor that holds either literals or the NodeID of its
+-- inputs, making it unique so long as NodeID's are unique
+data Node = NAdd NodeID NodeID
+          | NVariable String
+          | NConstant Int
+ deriving (Show,Eq,Ord)
+
+
+-- | A Directed Acyclic Graph is inside the values of the Trie
+data DAG = DAG { dagTrie :: Trie (Node,NodeID) -- | a trie from bytestrings to Node/NodeID
+               , dagMaxID :: Int -- | used to track # of hashcons performed
+               } deriving Show
+
+-- | DAG construction representation via the State monad
+data Graph a = Graph { unGraph :: State DAG NodeID
+                     , unStringAST :: ByteString }
+
+-- | Build a AST encoded as a ByteString
+buildStringAST :: Node -> [ByteString] -> ByteString
+buildStringAST node args =
+  let
+    opString = case node of
+                 NAdd _ _ -> "nadd"
+                 NVariable s -> "$" <> ByteString.pack s
+                 NConstant c -> "#" <> (ByteString.pack $ show c)
+    argsString = "(" <> ByteString.concat (List.intersperse "," args) <> ")"
+  in opString <> argsString
+
+
+-- TODO call me something different then hashcons?
+hashcons :: ByteString -> Node -> State DAG NodeID
+hashcons sAST node = do
+ DAG trie maxID <- get
+ case Trie.lookup sAST trie of
+   Nothing -> let maxID' = maxID+1
+                  trie' = Trie.insert sAST (node,maxID+1) trie
[…]

### 2022-12-15 paper/paper.tex
finished core body, needs polishing, abstract, intro, conclusion

@@
 \section{Implicit Sharing Via ByteString ASTs}
-TODO
+
+The heart of our problem is whenever we need to sequence the state of the inputs
+for one of our DSL functions we want to first check if it's already been
+evaluated. But how do we do that without first evaluating it to gain access to
+it's unique NodeID. We need some other way to uniquely identify it.
+
+Our proposed solution is too build an AST using byte strings along with our
+DAG, but hold it outside of the State monad. We can then build our DAG using a
+Trie, using the byte string AST to lookup our node identifiers instead.
+% TODO reference literature on tries
+
+\begin{minted}{haskell}
+data Graph a = Graph { unGraph :: State DAG NodeID
+                     , stringAST :: ByteString }
+
+data DAG = DAG { Trie (Node,NodeID)
+               } deriving Show
+\end{minted}
+
+The Trie essentially serves as our new BiMap, the resulting DAG contained in its
+lookup values. We essentially still perform the hash-consing technique but using
+the AST to perform the lookup
+\begin{minted}{haskell}
+hashcons :: ByteString -> Node -> State DAG NodeID
+hashcons sAST node = do
+ DAG trie maxID <- get
+ case Trie.lookup sAST trie of
+   Nothing -> let maxID' = maxID+1
+                  trie' = Trie.insert sAST (node,maxID+1) trie
+               in do put $ DAG trie' maxID'
+                     return maxID'
+   Just (_,nodeID) -> return nodeID
+
+instance Exp Graph where
+  constant x = let
+    node = NConstant x
+    sAST = buildStringAST node []
+    in Graph (hashcons sAST $ NConstant x) sAST
+  variable x = let
+    node = NVariable x
+    sAST = buildStringAST node []
+    in Graph (hashcons sAST $ NVariable x) sAST
+  add e1 e2 = let
+      sAST = buildStringAST (NAdd undefined undefined) [unStringAST e1,unStringAST e2]
+      sT = do ns <- seqArgs [e1,e2]
+              case ns of
+                [n1,n2] -> hashcons sAST $ NAdd n1 n2
+                _ -> error "black magic"
+    in Graph sT sAST
+
+\end{minted}
+
+The instance implementations for \mintinline{haskell}{constant} and
+\mintinline{haskell}{variable} work roughly the same, the novelty of the method
+is in how we handle DSL functions that take other DSL State as input like
+\mintinline{haskell}{add}. First we need to construct a byte string AST from
+it's input ASTs, there's a lot of ways we could go about this to attempt to
+minimize memory. A naive implementation would look similar to a pretty printer.
+Then we need to sequence it's inputs without evaluating the inner state if
+unnecessary. We do this through the implementation of \mintinline{haskell}{seqArgs}
+\begin{minted}{haskell}
+seqArgs :: [Graph a] -> State DAG [NodeID]
+seqArgs inps =
+  let
+    seqArg (Graph sT sAST) =
+      do DAG trie _ <- get
+         case Trie.lookup sAST  trie of
+           Nothing -> sT
+           Just (_,nodeID) -> return nodeID
+  in sequence $ map seqArg inps
+\end{minted}
+
+We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
+fail to look up its corresponding byte string AST in the Trie. This will prevent
+redundant hashconsing without the need for explicit sharing. However this method
+suffers from its own drawbacks.
 
 \subsection{Memory Limitations}
-TODO
+The byte string AST being built will itself suffer from lack of sharing. We're
+essentially trading extra computation for extra memory. This is often a good
+tradeoff, since memory is so plentiful in modern hardware. But under the right
+conditions it can become an issue
+
+TODO include heap profiling analysis
+
 \section{Explicit Sharing Of ByteString ASTs}
+We propose another solution to this issue, taking inspiration again from the
+\cite{kiselyov:sharing}, we can introduce an explicit construct for specifying
+sharing. This time, the construct will substitute the current byte string for a
+more compact label. For safety purposes, we need to keep track of a table of
+these labels and their corresponding ASTs, to make sure we don't insert of the
+same labels.
+\begin{minted}{haskell}
+data DAG = DAG { dagTrie :: Trie (Node,NodeID)
+               , dagCacheMap :: Map ByteString ByteString
+               } deriving Show
+
+data Graph a = Graph { unGraph :: State DAG NodeID
+                     , stringAST :: ByteString
+                     , addCache :: Maybe ByteString }
+
+class Cacheable repr where
+  cache :: ByteString -> repr ByteString -> repr ByteString
+instance Cacheable Graph where
+  cache s' (Graph g s _) = Graph g s' (Just s)
+
+-- TODO better example?
+test x y = let
+  z = cache "z" (add x y)
+  in add z z
+\end{minted}
+
+The cache operation replaces the current byte string AST with a new label, and
+we'll define a new operation \mintinline{haskell}{runCache} that will check if
+the label already exists in the cache map before inserting it.
+\begin{minted}{haskell}
+runCache :: ByteString -> Maybe ByteString -> Trie ByteString -> State DAG ()
+runCache sAST mAddCache cacheMap = do
+  case mAddCache of
+    Nothing -> return ()
+    Just sAST0 ->
+      case Trie.lookup sAST cacheMap of
+        Nothing -> let cacheMap' = Trie.insert sAST sAST0 cacheMap
+                   in modify (\dag -> dag { dagCache = cacheMap' })
+        Just sAST1 -> if sAST1 == sAST0
+                         then return ()
+                         else error $ "attempted to recache: " ++ show sAST
+
+seqArgs :: [Graph a] -> State DAG [NodeID]
+seqArgs inps =
+  let
+    seqArg (Graph sT sAST mCache) =
+      do DAG trie cacheMap _ <- get
+         runCache sAST mCache cacheMap
+         ...
+\end{minted}
+We need to make sure we don't attempt to insert the same cache label for two
+different ASTs. Unfortunately, if there is a collision there's no way to escape
+the State monad to prevent or modify the subtitution. The best we can do is
+crash the program, or if we use a monad tansformer, we ould use
+Control.Monad.Except to through an exception. Either way it's up to the DSL
+writer to insure they don't reuse the same label.
+
+\section{Conclusion}
 TODO
@@
 
-% LocalWords:  DSLs unsafePerformIO Haskell's AST typeclass tagless
+% LocalWords:  DSLs unsafePerformIO Haskell's AST typeclass tagless Trie
 % Local Variables:

### 2022-12-15 AltSharing/src/ByteStringSharing.hs

@@ -92,3 +92,7 @@ instance Exp Graph where
 
 
 buildDAG g = runState (unGraph g) (DAG Trie.empty 0)
+
+-- | Example of exponential scale of hash-consing
+addChains :: Exp repr => Int -> repr Int -> repr Int
+addChains n x0 = head $ drop n $ iterate (\x -> add x x) x0

### 2022-12-15 paper/paper.tex
switched from minted to lstlisting

@@
 \usepackage{listings}
-\usepackage{minted}
+% \usepackage{lstlisting}
+
 % Used for displaying a sample figure. If possible, figure files should
@@
 
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 data Exp
@@
 exp1 = Add exp0 exp0
-\end{minted}
+\end{lstlisting}
 
@@
 
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 class Exp repr where
@@
   constant :: Int -> repr Int
-\end{minted}
+\end{lstlisting}
 
@@
 For example, we can generate the AST from the previous DSL like so
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 newtype ExpR a = ExpR { unExpR :: Exp }
@@
   add (ExpR x) (ExpR y) = ExpR (Add x y)
-\end{minted}
+\end{lstlisting}
 
 Or we can implement pretty printing
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 newtype Pretty a = Pretty { unPretty :: String }
@@
   constant x = Pretty $ show x
-\end{minted}
+\end{lstlisting}
 
 And use the same DSL code to run either implementation
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 exp :: Exp repr => repr Int -> repr Int
@@
 expP = unPretty $ exp $ variable "v0"
-\end{minted}
+\end{lstlisting}
 
@@
 method first involves defining a DAG type, for example
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 type NodeID = Int
@@
 newtype DAG = DAG (BiMap Node) deriving Show
-\end{minted}
+\end{lstlisting}
 
@@
 
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 newtype Graph a = Graph { unGraph :: State DAG NodeID }
@@
                      hashcons $ NAdd h1 h2)
-\end{minted}
+\end{lstlisting}
 
@@
 before checking if it is already there.
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 hashcons :: Node -> State DAG NodeID
@@
     Just k -> return k
-\end{minted}
+\end{lstlisting}
 
@@
 once
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 haskellSharing x y=
@@
  in z + z
-\end{minted}
+\end{lstlisting}
 
@@
 equivalent attempt at using Haskell's built-in sharing in the finally tagless DSL
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 dslSharing :: Exp Graph -> Exp Graph -> Exp Graph
@@
   in add z z
-\end{minted}
+\end{lstlisting}
 Note {\bf z} is a wrapper around a State monad. Recall the implementation of
 {\bf add} via hash consing
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 add e1 e2 = Graph (do
@@
                    hashcons $ NAdd h1 h2)
-\end{minted}
+\end{lstlisting}
 The values {\bf h1} and {\bf h2} need to be explicitely evaluated through the
@@
 Consider a chain of {\bf add}'s  with sharing, for example
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 addChains :: Exp repr => Expr Int -> Expr Int
@@
   in xn
-\end{minted}
+\end{lstlisting}
 
@@
 custom let construct
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 class ExpLet repr where
@@
   )))
-\end{minted}
+\end{lstlisting}
 This makes the code a bit clunky and adds an extra burden on the DSL writer, but
@@
 outputs, such as tuples or container types like lists (for example
-\mintinline{haskell}{vadd :: (repr Int,repr Int) -> (repr Int,repr Int)}). %TODO
+\lstinline{haskell}{vadd :: (repr Int,repr Int) -> (repr Int,repr Int)}). %TODO
                                 % need to give implementation of this function?
@@
 correspond to the number of outputs
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 class ExpLet repr where
@@
        in (g0,g1)
-\end{minted}
+\end{lstlisting}
 If need be, its possible to enumerate custom let instances for every amount of
@@
 
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 data Graph a = Graph { unGraph :: State DAG NodeID
@@
                } deriving Show
-\end{minted}
+\end{lstlisting}
 
@@
 the AST to perform the lookup
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 hashcons :: ByteString -> Node -> State DAG NodeID
@@
 
-\end{minted}
+\end{lstlisting}
 
-The instance implementations for \mintinline{haskell}{constant} and
-\mintinline{haskell}{variable} work roughly the same, the novelty of the method
+The instance implementations for \lstinline{haskell}{constant} and
+\lstinline{haskell}{variable} work roughly the same, the novelty of the method
 is in how we handle DSL functions that take other DSL State as input like
-\mintinline{haskell}{add}. First we need to construct a byte string AST from
+\lstinline{haskell}{add}. First we need to construct a byte string AST from
 it's input ASTs, there's a lot of ways we could go about this to attempt to
@@
 Then we need to sequence it's inputs without evaluating the inner state if
-unnecessary. We do this through the implementation of \mintinline{haskell}{seqArgs}
-\begin{minted}{haskell}
+unnecessary. We do this through the implementation of \lstinline{haskell}{seqArgs}
+\begin{lstlisting}{haskell}
 seqArgs :: [Graph a] -> State DAG [NodeID]
@@
   in sequence $ map seqArg inps
-\end{minted}
+\end{lstlisting}
 
-We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
+We only evaluate the inner state \lstinline{haskell}{sT} of each argument if we
 fail to look up its corresponding byte string AST in the Trie. This will prevent
@@
 same labels.
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 data DAG = DAG { dagTrie :: Trie (Node,NodeID)
@@
   in add z z
-\end{minted}
+\end{lstlisting}
 
 The cache operation replaces the current byte string AST with a new label, and
-we'll define a new operation \mintinline{haskell}{runCache} that will check if
+we'll define a new operation \lstinline{haskell}{runCache} that will check if
 the label already exists in the cache map before inserting it.
-\begin{minted}{haskell}
+\begin{lstlisting}{haskell}
 runCache :: ByteString -> Maybe ByteString -> Trie ByteString -> State DAG ()
@@
          ...
-\end{minted}
+\end{lstlisting}
 We need to make sure we don't attempt to insert the same cache label for two
@@
 % Local Variables:
-% LaTeX-verbatim-environments-local: ("minted")
-% eval: (setq-local LaTeX-indent-environment-list (cons '("minted" current-indentation) (default-value 'LaTeX-indent-environment-list)))
+% LaTeX-verbatim-environments-local: ("lstlisting")
+% eval: (setq-local LaTeX-indent-environment-list (cons '("lstlisting" current-indentation) (default-value 'LaTeX-indent-environment-list)))
 % End:

### 2022-12-15 paper/paper.tex
just missing abstract and memory profiling

--- 2022-12-15 paper/paper.tex
@@
 \usepackage{minted}
+
 % Used for displaying a sample figure. If possible, figure files should
@@
 
-TODO describe sharing problem (mention observable sharing
-\cite{gill:observablesharing} and implicit/explicit sharing
-\cite{kiselyov:sharing} papers)
+In his paper ``Implementing explicit and finding implicit sharing in embedded
+DSLs'' \cite{kiselyov:sharing}, Kiselyov presents a method for implementing
+eDSLs in finally tagless form that generate a directed acyclic graph (DAG) with
+sharing. The interface for such DSL's uses the following format
+\begin{minted}{haskell}
+class Exp repr where
+  variable :: String -> repr Int
+  constant :: String -> repr Int
+  add :: repr Int -> repr Int -> repr Int
+  novel :: (repr Int,repr Int) -> (repr Int,repr Int)
+\end{minted}
+However, as we will explain in sections~\ref{limithashcons} and
+\ref{limitexplicit}, when the DSL contains functions which return multiple
+outputs (e.g., tuples, lists, etc.) such as \mintinline{haskell}{novel} in the
+above code snippet, Kiselyov's method of detecting sharing scales exponentially,
+and their method of explicitly declaring sharing is inapplicable. In our work,
+this translated into the inability to process large library functions.
 
-TODO describe finally tagless \cite{carette:finallytagless}
+In this paper, we will review Kiselyov's methods, identifying core issue, and
+present methods for implementing embedded DSLs with sharing that are both safe
+and maintain all the benefits of being embedded in the Haskell ecosystem. This
+means DSL functions are type-safe, do not require the use of unsafe referencing
+(i.e., via unsafePerformIO) and can return Haskell's container types (i.e.,
+tuples, lists, etc) without breaking sharing.
 
-We present methods for implementing embedded DSLs with sharing that are both
-safe and maintain all the benefits of being embedded in the Haskell ecosystem.
-This means DSL functions are type-safe, do not require the use of unsafe
-referencing (i.e., via unsafePerformIO) and can return Haskell's container
-types (i.e., tuples, lists, etc) without breaking sharing.
-
-\section{Detecting Sharing}
+\section{Background: Detecting Sharing}
 
@@
 -- Example
-v0 = Variable ``v0''
+v0 = Variable "v0"
 exp0 = Add v0 (Constant 0)
@@
 
-TODO describe DAG conversion by pointer comparison in the state monad
-
-\subsection{Detecting Sharing In Finally Tagless DSLs}
+TODO describe sharing through pointer comparison in monads
+\subsection{Finally Tagless DSLs}
 
@@
 We can then create different instances to implement different functionality.
-For example, we can generate the AST from the previous DSL like so
-\begin{minted}{haskell}
-newtype ExpR a = ExpR { unExpR :: Exp }
-
-instance Exp ExpR where
-  constant = ExpR . Constant
-  variable = ExpR . Variable
-  add (ExpR x) (ExpR y) = ExpR (Add x y)
-\end{minted}
-
-Or we can implement pretty printing
+For example, we can implement a pretty printer for our AST like so
 \begin{minted}{haskell}
@@
 
-And use the same DSL code to run either implementation
-\begin{minted}{haskell}
-exp :: Exp repr => repr Int -> repr Int
-exp v0 =
-  let
-    exp0 = add v0 (constant 0)
-  in add exp0 exp0
-
-expR = unExpR $ exp $ variable "v0"
-expP = unPretty $ exp $ variable "v0"
-\end{minted}
-
 Finally tagless style provides extensible, user friendly DSLs. However there
@@
 \subsection{Implicit Sharing Via Hash-Consing}
-TODO
-cite Ershov’s original description of hash-consing \cite{ershov1958:consing}
-cite Type safe consing implementation (with performance benchmarks) \cite{filliatre:typesafeconsing}
 
-In \cite{kiselyov:sharing}, a solution for detection implicit sharing in finally
-tagless style is presented via the method of hash-consing. You can find a more
-throrough explanation of the method there, but we'll give an overview here. This
-method first involves defining a DAG type, for example
+In this section, we'll provide an overview of Kiselyov's method in
+\cite{kiselyov:sharing} for detection of implicit sharing in finally tagless
+style is presented via the method of hash-consing, and discuss it's limitations
+in more detail. This method first involves defining a DAG type, for example
 \begin{minted}{haskell}
@@
 Note the purpose of the BiMap type is to be able to quickly insert and lookup
-nodes by their NodeID (i.e., a bijection of Node's and their NodeID's), and is
-most optimally implemented as a hash table with linear probing. The
-representation for the finally tagless instance is then a wrapper around a State
-monad that holds DAG in its state and returns the current (top) NodeID.
+nodes (i.e., a bijection of node's and their unique identifiers), and is most
+optimally implemented as a hash table with linear probing. Such an
+implementation of hash-consing using linear probing for an embedded DSL is given
+in thesis \cite{thai2021type}. The representation for the finally tagless
+instance is then a wrapper around a State monad that holds DAG in its state and
+returns the current (top) NodeID.
 
@@
 \end{minted}
+The technique is essentially that of
+hash-consing, popularized by its use in LISP compilers, but discovered long
+before that in 1958 \cite{ershov1958:consing}. Other works have explored the use
+of type safe hash consing in embedded DSLs, see \cite{filliatre:typesafeconsing}.
 
-\subsection{Limitations of Hash-Consing}
+\subsection{Limitations of Hash-Consing} \label{limithashcons}
 
@@
 Note {\bf z} is a wrapper around a State monad. Recall the implementation of
-{\bf add} via hash consing
-\begin{minted}{haskell}
-add e1 e2 = Graph (do
-                   h1 <- unGraph e1
-                   h2 <- unGraph e2
-                   hashcons $ NAdd h1 h2)
-\end{minted}
-The values {\bf h1} and {\bf h2} need to be explicitely evaluated through the
-State monad, meaning even if {\bf e1} and {\bf e2} are the same shared Haskell
-value, their underlying computations will be performed twice. Hash-consing will
-prevent these redundencies from appearing in the resulting DAG, but the entire
-unshared AST wil still be traversed, performing a hash-cons on each node.
+{\bf add} via hash consing, the values {\bf h1} and {\bf h2} need to be
+explicitely evaluated through the State monad, meaning even if {\bf e1} and {\bf
+  e2} are the same shared Haskell value, their underlying computations will be
+performed twice. Hash-consing will prevent these redundencies from appearing in
+the resulting DAG, but the entire unshared AST wil still be traversed,
+performing a hash-cons on each node.
 
@@
   \includegraphics[width=0.6\textwidth]{figs/hashcons.png}
-  \caption{Number of hashcons to add operations performed} \label{fig:hashcons}
+  \caption{Number of calls to \mintinline{haskell}{hashcons} to \mintinline{haskell}{add} operations performed} \label{fig:hashcons}
 \end{figure}
@@
 
-\subsection{Explicit Sharing and Limitations}
+\subsection{Explicit Sharing and Limitations} \label{limitexplicit}
 
@@
 This makes the code a bit clunky and adds an extra burden on the DSL writer, but
-it prevents unnecessary hash-consing in our example. However the method has it's
-limitations, suppose we want to write a DSL function that returns multiple
+it prevents unnecessary hash-consing in our example.
+
+However the method has it's limitations, suppose we want to write a DSL function
+that returns multiple
 outputs, such as tuples or container types like lists (for example
-\mintinline{haskell}{vadd :: (repr Int,repr Int) -> (repr Int,repr Int)}). %TODO
-                                % need to give implementation of this function?
-First of all, we need to implement different versions of the custom let construct to
-correspond to the number of outputs
+\mintinline{haskell}{novel :: (repr Int,repr Int) -> (repr Int,repr Int)}). When
+we go to implement DAG generation, we'll be forced to split the State monad into
+two
 \begin{minted}{haskell}
-class ExpLet repr where
-  let_ :: repr a -> (repr a -> repr b) -> repr b
-  let_ 2 :: repr a -> (repr a -> (repr b,repr c)) -> (repr b,repr c)
-
-instance ExpLet repr where
+instance Exp Graph where
   ...
-   let_2 e f = let
-       g0 = Graph (do x <- unGraph e
-                        let (o0,o1) = f (Graph (return x))
-                        unGraph o0)
-       g1 = Graph (do x <- unGraph e
-                        let (o0,o1) = f (Graph (return x))
-                        unGraph o1)
-       in (g0,g1)
+  novel e1 e2 = let
+     g1 = Graph (do h1 <- unGraph e1
+                    h2 <- unGraph e2
+                    h3 <- hashcons $ Novel h1 h2
+                    hashcons $ NovelOut1 h3)
+     g2 = Graph (do h1 <- unGraph e1
+                    h2 <- unGraph e2
+                    h3 <- hashcons $ Novel h1 h2
+                    hashcons $ NovelOut2 h3)
+     in (g1,g2)
 \end{minted}
-If need be, its possible to enumerate custom let instances for every amount of
-outputs or container types we would need, we could even use template Haskell to
-accomplish this. However, this custom let construct now has a new source of
-redundancy in its outputs. Each output it returns will now have to individually
-evaluate it's input, so a chain of DSL functions that output 2 or more values
-will suffer from the same exponential scaling of hashcons.
+Each output it returns will now have to be individually evaluated, so a
+chain of DSL functions that output 2 or more values will suffer from the same
+exponential scaling of hashcons operation. Furthermore, there's no good way to
+explicitely define sharing via a custom let construct because of closure.
 
@@
   add e1 e2 = let
-      sAST = buildStringAST (NAdd undefined undefined) [unStringAST e1,unStringAST e2]
+      sAST = buildStringAST "nadd" [e1,e2]
       sT = do ns <- seqArgs [e1,e2]
@@
 
--- TODO better example?
 test x y = let
@@
 \begin{minted}{haskell}
-runCache :: ByteString -> Maybe ByteString -> Trie ByteString -> State DAG ()
 runCache sAST mAddCache cacheMap = do
@@
       case Trie.lookup sAST cacheMap of
-        Nothing -> let cacheMap' = Trie.insert sAST sAST0 cacheMap
+        Nothing -> let cacheMap' = insert sAST sAST0 cacheMap
                    in modify (\dag -> dag { dagCache = cacheMap' })
-        Just sAST1 -> if sAST1 == sAST0
+        Just sAST1 ->
+            if sAST1 == sAST0
                then return ()
@@
 
-seqArgs :: [Graph a] -> State DAG [NodeID]
 seqArgs inps =
@@
 \section{Conclusion}
-TODO
+We have presented a method for constructing finally tagless style DSL's with
+sharing detection, that allow for DSL functions to have multiple outputs via
+built-in haskell tuples of container types. It also avoids the use of unsafe
+referencing as performed when doing observable sharing like in the work of
+\cite{gill:observablesharing}.
 
-\subsubsection{Acknowledgements} Please place your acknowledgments at
-the end of the paper, preceded by an unnumbered run-in heading (i.e.
-3rd-level heading).
+The method has its drawbacks in terms of memory usage, but these can be overcome
+by explicitly specifying sharing. This does present an extra burden on the DSL
+writer to implement explicit sharing when necessary and ensure labels are not
+reused. Future work may investigate the use of a preprocessor or plugin to
+automate explicit sharing.
+
+\subsubsection{Acknowledgements} This work drew heavily upon the work of Oleg
+Kiselyov, who's work on implicit and explicit sharing and finally tagless styles
+are the basis of our method
 
@@
 % LocalWords:  DSLs unsafePerformIO Haskell's AST typeclass tagless Trie
+% LocalWords:  Kiselyov's haskell consing
 % Local Variables:

### 2022-12-15 AltSharing/src/ByteStringSharing.hs

@@ -32,7 +32,7 @@ data Node = NAdd NodeID NodeID
 
 -- | A Directed Acyclic Graph is inside the values of the Trie
 data DAG = DAG { dagTrie :: Trie (Node,NodeID) -- | a trie from bytestrings to Node/NodeID
-               , dagMaxID :: Int -- | used to track # of hashcons performed
+               , dagMaxID :: Int -- | used to track # of triecons performed
                } deriving Show
 
 -- | DAG construction representation via the State monad
@@ -51,9 +51,9 @@ buildStringAST node args =
   in opString <> argsString
 
 
--- TODO call me something different then hashcons?
-hashcons :: ByteString -> Node -> State DAG NodeID
-hashcons sAST node = do
+-- TODO call me something different then triecons?
+triecons :: ByteString -> Node -> State DAG NodeID
+triecons sAST node = do
  DAG trie maxID <- get
  case Trie.lookup sAST trie of
    Nothing -> let maxID' = maxID+1
@@ -68,7 +68,7 @@ seqArgs inps =
   let
     seqArg (Graph sT sAST) =
       do DAG trie _ <- get
-         case Trie.lookup sAST  trie of
+         case Trie.lookup sAST trie of
            Nothing -> sT
            Just (_,nodeID) -> return nodeID
   in sequence $ map seqArg inps
@@ -77,19 +77,18 @@ instance Exp Graph where
   constant x = let
     node = NConstant x
     sAST = buildStringAST node []
-    in Graph (hashcons sAST $ NConstant x) sAST
+    in Graph (triecons sAST $ NConstant x) sAST
   variable x = let
     node = NVariable x
     sAST = buildStringAST node []
-    in Graph (hashcons sAST $ NVariable x) sAST
+    in Graph (triecons sAST $ NVariable x) sAST
   add e1 e2 = let
       sAST = buildStringAST (NAdd undefined undefined) [unStringAST e1,unStringAST e2]
       sT = do ns <- seqArgs [e1,e2]
               case ns of
-                [n1,n2] -> hashcons sAST $ NAdd n1 n2
+                [n1,n2] -> triecons sAST $ NAdd n1 n2
                 _ -> error "black magic"
     in Graph sT sAST
-
 
 buildDAG g = runState (unGraph g) (DAG Trie.empty 0)
 

### 2022-12-15 AltSharing/src/HashConsSharing.hs

@@ -124,7 +124,6 @@ hashcons e = do
                   >> return k
     Just k -> return k
 
-
 -- | Example of exponential scale of hash-consing
 addChains :: Exp repr => Int -> repr Int -> repr Int
 addChains n x0 = head $ drop n $ iterate (\x -> add x x) x0

### 2022-12-16 paper/paper.tex
alpha read through

@@
 \author{Curtis D'Alves \and
-Christopher Anand \and
 Lucas Dutton \and
-Steven Gonder
+Steven Gonder \and
+Christopher Kumar Anand
 }
@@
 
-\keywords{First keyword  \and Second keyword \and Another keyword.}
+\keywords{DSL  \and sharing \and common-subexpression elimination \and Haskell.}
 \end{abstract}
@@
 
-In his paper ``Implementing explicit and finding implicit sharing in embedded
-DSLs'' \cite{kiselyov:sharing}, Kiselyov presents a method for implementing
+Kiselyov \cite{kiselyov:sharing}  presents a method for implementing
 eDSLs in finally tagless form that generate a directed acyclic graph (DAG) with
-sharing. The interface for such DSL's uses the following format
+sharing. 
+However, as we will explain in sections~\ref{limithashcons} and
+\ref{limitexplicit}, for DSL functions that return multiple
+outputs (e.g., tuples, lists, etc.), Kiselyov's method of detecting sharing may require computation exponential in the size of the program,
+and his method of explicitly declaring sharing is inapplicable.
+
+In the toy example
 \begin{minted}{haskell}
@@
 \end{minted}
-However, as we will explain in sections~\ref{limithashcons} and
-\ref{limitexplicit}, when the DSL contains functions which return multiple
-outputs (e.g., tuples, lists, etc.) such as \mintinline{haskell}{novel} in the
-above code snippet, Kiselyov's method of detecting sharing scales exponentially,
-and their method of explicitly declaring sharing is inapplicable. In our work,
+the function \mintinline{haskell}{novel} exhibits this problem.
+ In our work,
 this translated into the inability to process large library functions.
 
-In this paper, we will review Kiselyov's methods, identifying core issue, and
-present methods for implementing embedded DSLs with sharing that are both safe
-and maintain all the benefits of being embedded in the Haskell ecosystem. This
-means DSL functions are type-safe, do not require the use of unsafe referencing
-(i.e., via unsafePerformIO) and can return Haskell's container types (i.e.,
-tuples, lists, etc) without breaking sharing.
+\smallskip
+In this paper, we review Kiselyov's methods, identifying the core issue, and
+present methods for implementing embedded DSLs with sharing that avoid  unsafe referencing (i.e., \mintinline{haskell}{unsafePerformIO}) \cite{gill:observablesharing}, maintain all the benefits of being embedded in the Haskell ecosystem and are computationally feasible.
+This
+means DSL functions are pure, type-safe and can return Haskell container types (i.e.,
+tuples, lists, etc.) without breaking sharing.
 
@@
 
-A naive DSL implementation of an expression in Haskell can be done via standard
-Haskell data types, for example:
-
+Consider the naive DSL implemented as a Haskell data type:
 \begin{minted}{haskell}
@@
   | Constant Int
-
--- Example
+\end{minted}
+Expressions generate Abstract Syntax Trees (ASTs),
+but consider this example,
+\begin{minted}{haskell}
 v0 = Variable "v0"
@@
 \end{minted}
+in which the expression  \mintinline{haskell}{exp0} is shared,
+and will therefore be stored once in memory.
+For large expressions with lots of sharing,
+this can make a substantial difference.
+However,
+one of the first things the developer will do is write a pretty printer.
+That recursive function will traverse the data structure as a tree,
+and pretty print \mintinline{haskell}{exp0} twice.
+This inefficiency is a problem for code generation,
+and naive traversal of the AST does the opposite of common-subexpression elimination performed by a good optimizing compiler.
+To avoid this,
+rather than representing the code as an AST, 
+we should use a DAG, retaining all of the sharing in the original DSL code.
 
-Note the DSL generates a tree, or to be more specific an Abstract Syntax Tree
-(AST). Common features a DSL implementer might implement would include code
-generation or pretty printing. Simple traversal of the AST for either of these
-operations would result in duplication, for example in the above code snippet
-the AST for {\bf exp0} will be traversed twice. For code generation in
-particular this would be problematic, in order to circumvent this problem in
-general we need to perform common subexpression elimination by converting the
-AST into a directed acyclic graph. % #TODO cite literature
+One way of maintaining sharing is by observable sharing (see Section 3 in \cite{kiselyov:sharing}).
+In Haskell, this requires a monadic interface.
+Monads are useful, but don't match the expectations of domain experts \cite{odonnell:embedding}.
 
-TODO describe sharing through pointer comparison in monads
 \subsection{Finally Tagless DSLs}
 
-Monads are useful, but don't make for a very user friendly DSL. It would be nice
+ It would be nice
 to make use of monadic state when we need it (i.e., for converting to a DAG)
-while hiding it behind a nice pure interface. The final tagless approach of
+while hiding it behind a nice pure interface. The finally tagless approach 
 \cite{carette:finallytagless} is popular for accomplishing this. In this
-approach, DSL expressions are built using typeclass methods that wrap the DSL in
-a parameterized representation. For example, the previous data type based DSL
-could be written in finally tagless as
+approach, DSL expressions are built using type-class methods that wrap the DSL in
+a parameterized representation. For example, the previous data-type-based DSL
+could be written in finally tagless style as
 
@@
 We can then create different instances to implement different functionality.
-For example, we can implement a pretty printer for our AST like so
+For example, we can implement a pretty printer for our AST as
 \begin{minted}{haskell}
@@
 
-Finally tagless style provides extensible, user friendly DSLs. However there
-are still some complications when using it to implement sharing.
+Finally tagless style provides extensible, user friendly DSLs.
 
-\subsection{Implicit Sharing Via Hash-Consing}
+\subsection{Implicit Sharing via Hash-Consing}
 
-In this section, we'll provide an overview of Kiselyov's method in
-\cite{kiselyov:sharing} for detection of implicit sharing in finally tagless
-style is presented via the method of hash-consing, and discuss it's limitations
-in more detail. This method first involves defining a DAG type, for example
+Kiselyov's method
+ for detecting implicit sharing in finally tagless
+style uses hash-consing \cite{kiselyov:sharing}. 
+Hash-consing is based on a bijection of nodes and a set of identifiers,
+e.g., with interface 
+\begin{minted}{haskell}
+data BiMap a -- abstract
+lookup_key :: Ord a => a -> BiMap a -> Maybe Int
+lookup_val :: Int -> BiMap a -> a
+insert :: Ord a => a -> BiMap a -> (Int,BiMap a)
+empty :: BiMap a
+\end{minted}
+An efficient implementation using hashing and linear probing is given by Thai
+in his Master's thesis \cite{thai2021type}.
+
+Nodes need to be uniquely identifiable, and shouldn't be a recursive data type,
+such as
 \begin{minted}{haskell}
@@
           | NConstant Int
+\end{minted}
+The representation for the finally tagless
+instance is then a wrapper around a State monad that holds the DAG being constructed in its state and
+returns the current (top) \mintinline{haskell}{NodeID}:
+\begin{minted}{haskell}
+newtype DAG = DAG (BiMap Node) deriving Show
 
-data BiMap a -- abstract
-lookup_key :: Ord a => a -> BiMap a -> Maybe Int
-lookup_val :: Int -> BiMap a -> a
-insert :: Ord a => a -> BiMap a -> (Int,BiMap a)
-empty :: BiMap a
-
-newtype DAG = DAG (BiMap Node) deriving Show
-\end{minted}
-
-Note the purpose of the BiMap type is to be able to quickly insert and lookup
-nodes (i.e., a bijection of node's and their unique identifiers), and is most
-optimally implemented as a hash table with linear probing. Such an
-implementation of hash-consing using linear probing for an embedded DSL is given
-in thesis \cite{thai2021type}. The representation for the finally tagless
-instance is then a wrapper around a State monad that holds DAG in its state and
-returns the current (top) NodeID.
-
-\begin{minted}{haskell}
 newtype Graph a = Graph { unGraph :: State DAG NodeID }
@@
 \end{minted}
-
-The trick to uncovering sharing in the implementation is implemented via the
-{\bf hashcons} function, which inserts a new node into the current DAG, but not
+The trick to uncovering sharing is in the
+\mintinline{hashcons} function, which inserts a new node into the current DAG, but not
 before checking if it is already there.
@@
 The technique is essentially that of
-hash-consing, popularized by its use in LISP compilers, but discovered long
-before that in 1958 \cite{ershov1958:consing}. Other works have explored the use
-of type safe hash consing in embedded DSLs, see \cite{filliatre:typesafeconsing}.
+hash-consing, popularized by its use in LISP compilers, but discovered by Ershov in 1958 \cite{ershov1958:consing}. Other works have explored the use
+of type-safe hash-consing in embedded DSLs, see \cite{filliatre:typesafeconsing}.
 
@@
 
-When we wrap our State monad in finally tagless style, we lose some of Haskell's
-built-in sharing capability. Consider the following code, note that the use of
-local variables explicitely defines the computation $x + y$ to only be computed
-once
+When we wrap our State monad in finally tagless style, we lose some expected sharing. In the following code, the use of the let causes the computation $x + y$ to only occur once
 \begin{minted}{haskell}
@@
 \end{minted}
-Note {\bf z} is a wrapper around a State monad. Recall the implementation of
-{\bf add} via hash consing, the values {\bf h1} and {\bf h2} need to be
-explicitely evaluated through the State monad, meaning even if {\bf e1} and {\bf
-  e2} are the same shared Haskell value, their underlying computations will be
-performed twice. Hash-consing will prevent these redundencies from appearing in
-the resulting DAG, but the entire unshared AST wil still be traversed,
-performing a hash-cons on each node.
+Knowing that \mintinline{haskell}{z} is a wrapper around a state monad,
+and recalling the implementation of
+\mintinline{haskell}{add} via hash-consing above, 
+the values \mintinline{haskell}{h1} and \mintinline{haskell}{h2} are
+separately evaluated through the state monad, even if \mintinline{haskell}{e1} and \mintinline{haskell}{e2} are the same shared Haskell value. 
+Hash-consing will prevent these redundancies from appearing in
+the resulting DAG, 
+but in the process of discovering the sharing, the entire unshared AST will still be traversed.
 
-Consider a chain of {\bf add}'s  with sharing, for example
+Consider a chain of \mintinline{haskell}{add}s  with sharing, for example
 \begin{minted}{haskell}
 addChains :: Exp repr => Expr Int -> Expr Int
-addChains x0 = let
+addChains x0 = 
+  let
     x1 = add x0 x0
@@
   \includegraphics[width=0.6\textwidth]{figs/hashcons.png}
-  \caption{Number of calls to \mintinline{haskell}{hashcons} to \mintinline{haskell}{add} operations performed} \label{fig:hashcons}
+  \caption{Number of calls to \mintinline{haskell}{hashcons} plotted against the number of \mintinline{haskell}{add} operations performed.} \label{fig:hashcons}
 \end{figure}
-As you can see from figure \ref{fig:hashcons}, this code will perform approximately
-$2^{n+1}$ hashcons operations, where n is the number of {\bf add}'s.
+As shown in Fig.~\ref{fig:hashcons}, this code will perform approximately
+$2^{n+1}$ \mintinline{haskell}{hashcons} operations, where $n$ is the number of \mintinline{haskell}{add}s.
 
@@
 
-\cite{kiselyov:sharing} acknowledges the amount of computation with hash-consing
-can get out of control, and proposes an ad-hoc solution, explicit sharing via a
+Kiselyov \cite{kiselyov:sharing} recognized that the amount of computation with hash-consing ``may take a long time for large programs,'' 
+and proposed an ad-hoc solution, explicit sharing via a
 custom let construct
@@
                      unGraph $ f (Graph (return x)))
-
+\end{minted}
+which can be used to rewrite \mintinline{haskell}{addChains} as
+\begin{minted}{haskell}
 addChains x =
@@
 
-However the method has it's limitations, suppose we want to write a DSL function
-that returns multiple
-outputs, such as tuples or container types like lists (for example
-\mintinline{haskell}{novel :: (repr Int,repr Int) -> (repr Int,repr Int)}). When
-we go to implement DAG generation, we'll be forced to split the State monad into
-two
+However the method does not work for DSL functions
+returning multiple outputs via tuples or container types like lists. Recall the definition
+\begin{minted}{haskell}
+novel :: (repr Int,repr Int) -> (repr Int,repr Int)
+\end{minted}
+The problem is that DAG generation requires splitting the state monad in two:
 \begin{minted}{haskell}
@@
                     h2 <- unGraph e2
-                    h3 <- hashcons $ Novel h1 h2
-                    hashcons $ NovelOut1 h3)
+                    hashcons $ Novel1 h1 h2)
      g2 = Graph (do h1 <- unGraph e1
                     h2 <- unGraph e2
-                    h3 <- hashcons $ Novel h1 h2
-                    hashcons $ NovelOut2 h3)
+                    hashcons $ Novel2 h1 h2)
      in (g1,g2)
@@
 chain of DSL functions that output 2 or more values will suffer from the same
-exponential scaling of hashcons operation. Furthermore, there's no good way to
-explicitely define sharing via a custom let construct because of closure.
+exponential explosion of hashcons operations,
+and trying to adapt the let construct above, 
+just creates another function function with the same problem (multiple outputs).
 
 One solution to this issue is to integrate container types such as tuples and
-lists into the DSL language. However doing this will take away form the
-advantages of having an embedded language, manipulating tuple values will be
-cumbersome constantly requiring calls to custom implementations of {\bf
-  fst}/{\bf snd} etc. And for lists you'll lose all access to built-in Haskell
+lists into the DSL language. However doing this eliminates the
+advantage of having an embedded language.
+Manipulating tuple values will be
+cumbersome, constantly requiring calls to custom implementations of \mintinline{haskell}{fst},\mintinline{haskell}{snd} etc. 
+And for lists you'll lose access to built-in Haskell
 list functionality.
 
-\section{Implicit Sharing Via ByteString ASTs}
+\section{Implicit Sharing Via Byte String ASTs}
 
-The heart of our problem is whenever we need to sequence the state of the inputs
+The heart of our problem is that whenever we need to sequence the state of the inputs
 for one of our DSL functions we want to first check if it's already been
 evaluated. But how do we do that without first evaluating it to gain access to
-it's unique NodeID. We need some other way to uniquely identify it.
+its unique identifier. We need some way to uniquely identify it outside the monad.
 
-Our proposed solution is too build an AST using byte strings along with our
-DAG, but hold it outside of the State monad. We can then build our DAG using a
-Trie, using the byte string AST to lookup our node identifiers instead.
+Our proposed solution is to build a serialized AST using byte strings for each node along with our
+DAG.
+The byte string stays outside monad, while the DAG remains inside.
+We can do this efficiently by replacing the \mintinline{haskell}{BiMap}
+with a trie.
+In our toy example, we use package \texttt{bytestring-trie}.
 % TODO reference literature on tries
@@
 
-data DAG = DAG { Trie (Node,NodeID)
+data DAG = DAG { unTrie :: Trie (Node,NodeID)
+               , maxID  :: NodeID
                } deriving Show
 \end{minted}
+This looks a bit different because the \mintinline{haskel}{BiMap}
+was a bijective relation between nodes and node ids,
+whereas the trie maps byte strings to pairs (node,node id).
+To get the DAG expressed as a relation, project out the values of the trie.
 
-The Trie essentially serves as our new BiMap, the resulting DAG contained in its
-lookup values. We essentially still perform the hash-consing technique but using
-the AST to perform the lookup
+To prevent confusion, we rename the hash-consing function to \mintinline{haskel}{triecons}, as follows
 \begin{minted}{haskell}
-hashcons :: ByteString -> Node -> State DAG NodeID
-hashcons sAST node = do
+triecons :: ByteString -> Node -> State DAG NodeID
+triecons sAST node = do
   DAG trie maxID <- get
@@
     Nothing -> let maxID' = maxID+1
-                  trie' = Trie.insert sAST (node,maxID+1) trie
+                   trie' = Trie.insert sAST (node,maxID') trie
                 in do put $ DAG trie' maxID'
@@
     Just (_,nodeID) -> return nodeID
-
+\end{minted}
+We use it to implement the DAG-building instance of the DSL,
+which looks a lot like the previous instance.
+The substantial differences are the 
+\mintinline{haskell}{buildStringAST} calls
+which you can think of as pretty printing, but optimized for the trie,
+and the use of \mintinline{haskell}{seqArgs} (explained below):
+\begin{minted}{haskell}
 instance Exp Graph where
@@
     sAST = buildStringAST node []
-    in Graph (hashcons sAST $ NConstant x) sAST
+    in Graph (triecons sAST $ NConstant x) sAST
   variable x = let
@@
     sAST = buildStringAST node []
-    in Graph (hashcons sAST $ NVariable x) sAST
+    in Graph (triecons sAST $ NVariable x) sAST
   add e1 e2 = let
@@
               case ns of
-                [n1,n2] -> hashcons sAST $ NAdd n1 n2
+                [n1,n2] -> triecons sAST $ NAdd n1 n2
                 _ -> error "black magic"
     in Graph sT sAST
-
 \end{minted}
-
-The instance implementations for \mintinline{haskell}{constant} and
-\mintinline{haskell}{variable} work roughly the same, the novelty of the method
-is in how we handle DSL functions that take other DSL State as input like
-\mintinline{haskell}{add}. First we need to construct a byte string AST from
-it's input ASTs, there's a lot of ways we could go about this to attempt to
-minimize memory. A naive implementation would look similar to a pretty printer.
-Then we need to sequence it's inputs without evaluating the inner state if
-unnecessary. We do this through the implementation of \mintinline{haskell}{seqArgs}
+The magic is in \mintinline{haskell}{seqArgs}.
+We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
+fail to look up its corresponding byte string AST in the Trie.
 \begin{minted}{haskell}
@@
 
-We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
-fail to look up its corresponding byte string AST in the Trie. This will prevent
-redundant hashconsing without the need for explicit sharing. However this method
-suffers from its own drawbacks.
+This will prevent
+redundant hashconsing without the need for explicit sharing,
+but at the expense of storing redundant byte strings.
 
@@
 more compact label. For safety purposes, we need to keep track of a table of
-these labels and their corresponding ASTs, to make sure we don't insert of the
-same labels.
+these labels and their corresponding ASTs, to make sure we don't use the same label for different ASTs.
 \begin{minted}{haskell}
@@
                , dagCacheMap :: Map ByteString ByteString
+               , maxID  :: NodeID
                } deriving Show
@@
 The cache operation replaces the current byte string AST with a new label, and
-we'll define a new operation \mintinline{haskell}{runCache} that will check if
+we define a new operation \mintinline{haskell}{runCache} to check if
 the label already exists in the cache map before inserting it.
@@
     Just sAST0 ->
-      case Trie.lookup sAST cacheMap of
+      case Map.lookup sAST cacheMap of
         Nothing -> let cacheMap' = insert sAST sAST0 cacheMap
@@
 different ASTs. Unfortunately, if there is a collision there's no way to escape
-the State monad to prevent or modify the subtitution. The best we can do is
-crash the program, or if we use a monad tansformer, we ould use
-Control.Monad.Except to through an exception. Either way it's up to the DSL
+the State monad to prevent or modify the substitution. 
+In the toy example, compilation crashes, but we could catch an exception instead.
+Either way it's up to the DSL
 writer to insure they don't reuse the same label.
@@
 \section{Conclusion}
-We have presented a method for constructing finally tagless style DSL's with
-sharing detection, that allow for DSL functions to have multiple outputs via
-built-in haskell tuples of container types. It also avoids the use of unsafe
-referencing as performed when doing observable sharing like in the work of
-\cite{gill:observablesharing}.
+We have presented a method for constructing finally tagless style DSLs with
+sharing detection, that allow for multiple-output DSL functions. 
+It also avoids the use of unsafe referencing as performed when doing observable sharing, c.f. \cite{gill:observablesharing}.
 
-The method has its drawbacks in terms of memory usage, but these can be overcome
+The method has its drawbacks in terms of memory usage, but these can be mitigated
 by explicitly specifying sharing. This does present an extra burden on the DSL
@@
 
-\subsubsection{Acknowledgements} This work drew heavily upon the work of Oleg
-Kiselyov, who's work on implicit and explicit sharing and finally tagless styles
-are the basis of our method
-
+\subsubsection{Acknowledgements} We thank NSERC and IBM Canada for supporting this work.
 %

### 2022-12-16 paper/paper.tex
benchmarking toy problem

@@
 \author{Curtis D'Alves \and
+Christopher Kumar Anand \and
 Lucas Dutton \and
-Steven Gonder \and
-Christopher Kumar Anand
+Steven Gonder
 }
@@
 The trick to uncovering sharing is in the
-\mintinline{hashcons} function, which inserts a new node into the current DAG, but not
+\mintinline{haskell}{hashcons} function, which inserts a new node into the current DAG, but not
 before checking if it is already there.
@@
 with a trie.
-In our toy example, we use package \texttt{bytestring-trie}.
+In our toy example, we use the package \texttt{bytestring-trie}.
 % TODO reference literature on tries
@@
 \end{minted}
-This looks a bit different because the \mintinline{haskel}{BiMap}
+This looks a bit different because the \mintinline{haskell}{BiMap}
 was a bijective relation between nodes and node ids,
@@
 
-To prevent confusion, we rename the hash-consing function to \mintinline{haskel}{triecons}, as follows
+To prevent confusion, we rename the hash-consing function to \mintinline{haskell}{triecons}, as follows
 \begin{minted}{haskell}

### 2022-12-16 AltSharing/src/ByteStringExplicit.hs

@@ -21,7 +21,7 @@ class Exp repr where
 
 -- | Explicit caching of ByteStrings
 class Cacheable repr where
-  cache :: ByteString -> repr ByteString -> repr ByteString
+  cache :: ByteString -> repr a -> repr a
 
 instance Cacheable Graph where
   cache s' (Graph g s _) = Graph g s' (Just s)
@@ -39,7 +39,7 @@ data Node = NAdd NodeID NodeID
 
 -- | A Directed Acyclic Graph is inside the values of the Trie
 data DAG = DAG { dagTrie :: Trie (Node,NodeID) -- | a trie from bytestrings to Node/NodeID
-               , dagCache :: Trie ByteString -- | FIXME switch to Map?
+               , dagCache :: Map ByteString ByteString -- | FIXME switch to Map?
                , dagMaxID :: Int -- | used to track # of hashcons performed
                } deriving Show
 
@@ -82,13 +82,13 @@ seqArgs inps =
            Just (_,nodeID) -> return nodeID
   in sequence $ map seqArg inps
 
-runCache :: ByteString -> Maybe ByteString -> Trie ByteString -> State DAG ()
+runCache :: ByteString -> Maybe ByteString -> Map ByteString ByteString -> State DAG ()
 runCache sAST mCache cacheMap = do
   case mCache of
     Nothing -> return ()
     Just sAST0 ->
-      case Trie.lookup sAST cacheMap of
-        Nothing -> let cacheMap' = Trie.insert sAST sAST0 cacheMap
+      case Map.lookup sAST cacheMap of
+        Nothing -> let cacheMap' = Map.insert sAST sAST0 cacheMap
                    in modify (\dag -> dag { dagCache = cacheMap' })
         Just sAST1 -> if sAST1 == sAST0
                          then return ()
@@ -111,4 +111,12 @@ instance Exp Graph where
                 _ -> error "black magic"
     in Graph sT sAST Nothing
 
-buildDAG g = runState (unGraph g) (DAG Trie.empty Trie.empty 0)
+buildDAG g = runState (unGraph g) (DAG Trie.empty Map.empty 0)
+
+
+-- | Example of exponential scale of hash-consing
+addChains :: (Exp repr,Cacheable repr) => Int -> repr Int -> repr Int
+addChains n x0 = fst
+                 $ head
+                 $ drop n
+                 $ iterate (\(x,i) -> (cache ("add"<>(ByteString.pack $ show i)) (add x x),i+1)) (x0,0)

### 2022-12-16 AltSharing/src/ByteStringSharing.hs

@@ -13,6 +13,8 @@ import Data.Trie (Trie)
 import qualified Data.Trie as Trie
 import qualified Data.List as List
 
+import qualified HashConsSharing as HCS
+
 -- | Finally Tagless Style Expression DSL
 class Exp repr where
   add :: repr Int -> repr Int -> repr Int
@@ -32,7 +34,8 @@ data Node = NAdd NodeID NodeID
 
 -- | A Directed Acyclic Graph is inside the values of the Trie
 data DAG = DAG { dagTrie :: Trie (Node,NodeID) -- | a trie from bytestrings to Node/NodeID
-               , dagMaxID :: Int -- | used to track # of triecons performed
+               , dagNumCons :: Int -- | used to track number of triecons performed
+               , dagMaxID :: Int -- | used to track max node id
                } deriving Show
 
 -- | DAG construction representation via the State monad
@@ -54,11 +57,12 @@ buildStringAST node args =
 -- TODO call me something different then triecons?
 triecons :: ByteString -> Node -> State DAG NodeID
 triecons sAST node = do
- DAG trie maxID <- get
+ DAG trie tCnt maxID <- get
+ modify (\dag -> dag { dagNumCons = tCnt + 1})
  case Trie.lookup sAST trie of
    Nothing -> let maxID' = maxID+1
                   trie' = Trie.insert sAST (node,maxID+1) trie
-               in do put $ DAG trie' maxID'
+               in do modify (\dag -> dag { dagTrie = trie', dagMaxID = maxID' } )
                      return maxID'
    Just (_,nodeID) -> return nodeID
 
@@ -67,7 +71,7 @@ seqArgs :: [Graph a] -> State DAG [NodeID]
 seqArgs inps =
   let
     seqArg (Graph sT sAST) =
-      do DAG trie _ <- get
+      do DAG trie _ _ <- get
          case Trie.lookup sAST trie of
            Nothing -> sT
            Just (_,nodeID) -> return nodeID
@@ -90,8 +94,18 @@ instance Exp Graph where
                 _ -> error "black magic"
     in Graph sT sAST
 
-buildDAG g = runState (unGraph g) (DAG Trie.empty 0)
+buildDAG g = runState (unGraph g) (DAG Trie.empty 0 0)
 
 -- | Example of exponential scale of hash-consing
 addChains :: Exp repr => Int -> repr Int -> repr Int
[…]

### 2022-12-16 AltSharing/src/HashConsSharing.hs

@@ -96,7 +96,7 @@ empty = BiMap Map.empty 0
 
 -- | A Directed Acyclic Graph is a BiMap between Node and NodeID
 data DAG = DAG { dagMAP :: BiMap Node  -- | A map from Node to NodeID
-               , dagCnt :: Int -- | used to track # of hashcons performed
+               , dagNumCons :: Int -- | used to track # of hashcons performed
                } deriving Show
 
 -- | DAG construction representation via the State monad
@@ -117,7 +117,7 @@ buildDAG g = runState (unGraph g) (DAG empty 0)
 hashcons :: Node -> State DAG NodeID
 hashcons e = do
   DAG m cnt <- get
-  modify (\dag -> dag { dagCnt = cnt+1}) -- track number of hashcons performed
+  modify (\dag -> dag { dagNumCons = cnt+1}) -- track number of hashcons performed
   case lookup_key e m of
     Nothing -> let (k,m') = insert e m
                in modify (\dag -> dag { dagMAP = m' })
@@ -131,7 +131,7 @@ addChains n x0 = head $ drop n $ iterate (\x -> add x x) x0
 plotAddChains size =
   let
     chainsData = map (\(x,y) -> (fromIntegral x,fromIntegral y))
-      [ (n,dagCnt $ snd $ buildDAG $ addChains n $ variable "x") | n <- [0..size] ]
+      [ (n,dagNumCons $ snd $ buildDAG $ addChains n $ variable "x") | n <- [0..size] ]
   in plot (PNG "plot.png") $ Data2D [Title "Hashcons Scaling",Style Linespoints,Color Red] [] chainsData
 
 

### 2022-12-16 paper/paper.tex
fix in ByteStringExplicit and added profiling

@@
   \centering
-  \includegraphics[width=0.6\textwidth]{figs/hashcons.png}
+  \includegraphics[width=0.6\textwidth]{figs/hashconscmp.png}
   \caption{Number of calls to \mintinline{haskell}{hashcons} plotted against the number of \mintinline{haskell}{add} operations performed.} \label{fig:hashcons}
@@
 data DAG = DAG { dagTrie :: Trie (Node,NodeID)
-               , dagCacheMap :: Map ByteString ByteString
-               , maxID  :: NodeID
+               , dagSubMap :: Map ByteString ByteString
+               , dagMaxID :: Int
                } deriving Show
@@
 data Graph a = Graph { unGraph :: State DAG NodeID
-                     , stringAST :: ByteString
-                     , addCache :: Maybe ByteString }
+                     , unStringAST :: ByteString
+                     , unSubT :: Maybe ByteString }
 
-class Cacheable repr where
-  cache :: ByteString -> repr ByteString -> repr ByteString
-instance Cacheable Graph where
-  cache s' (Graph g s _) = Graph g s' (Just s)
+class Substitute repr where
+  subT :: ByteString -> repr a -> repr a
+instance Substitute Graph where
+  subT s' (Graph g s _) = Graph g s' (Just s)
 
 test x y = let
-  z = cache "z" (add x y)
+  z = subT "z" (add x y)
   in add z z
@@
 
-The cache operation replaces the current byte string AST with a new label, and
-we define a new operation \mintinline{haskell}{runCache} to check if
+The \mintinline{haskell}{subT} operation substitutes the current byte string AST with a new one, and
+we define a new operation \mintinline{haskell}{subTInsert} to check if
 the label already exists in the cache map before inserting it.
 \begin{minted}{haskell}
-runCache sAST mAddCache cacheMap = do
-  case mAddCache of
-    Nothing -> return ()
-    Just sAST0 ->
-      case Map.lookup sAST cacheMap of
-        Nothing -> let cacheMap' = insert sAST sAST0 cacheMap
-                   in modify (\dag -> dag { dagCache = cacheMap' })
-        Just sAST1 ->
-            if sAST1 == sAST0
-               then return ()
-               else error $ "attempted to recache: " ++ show sAST
-
+seqArgs :: [Graph a] -> State DAG [NodeID]
 seqArgs inps =
   let
-    seqArg (Graph sT sAST mCache) =
-      do DAG trie cacheMap _ <- get
-         runCache sAST mCache cacheMap
-         ...
+    seqArg (Graph sT sAST mSubt) =
+      do DAG trie _ _ <- get
+         let sAST' = case mSubt of
+                       Just s -> s
+                       Nothing -> sAST
+         case Trie.lookup sAST' trie of
+           Nothing -> sT -- error "missing ast"
+           Just (node,nodeID) ->
+              do subTInsert mSubt sAST (node,nodeID)
+                 return nodeID
+  in sequence $ map seqArg inps
+
+subTInsert :: Maybe ByteString -> ByteString
+           -> (Node, NodeID) -> State DAG ()
+subTInsert Nothing  _ _  = return ()
+subTInsert (Just s) sAST nodeID =
+  do DAG trie subtMap _ <- get
+     case Map.lookup sAST subtMap of
+        Just sAST' -> if sAST == sAST'
+                      then return ()
+                      else error "tried to resubT"
+        Nothing -> let cMap' = Map.insert sAST s subtMap
+                       trie' = Trie.insert sAST nodeID trie
+                  in modify (\dag -> dag { dagTrie = trie'
+                                      , dagSubMap = cMap' })
 \end{minted}
-We need to make sure we don't attempt to insert the same cache label for two
+We need to make sure we don't attempt to insert the same substitution for two
 different ASTs. Unfortunately, if there is a collision there's no way to escape

### 2022-12-16 AltSharing/app/Main.hs

@@ -1,6 +1,15 @@
 module Main where
 
-import Lib ()
+import qualified HashConsSharing as HCS
+import qualified ByteStringSharing as BSS
+import qualified ByteStringExplicit as BSE
 
+import qualified Data.Trie as Trie
+import qualified Data.Map as Map
+
+n = 2000
 main :: IO ()
-main = putStrLn "Hello World"
+main = let
+   -- dag = Map.toList $ HCS.biMap $ HCS.dagMAP $ snd $ HCS.buildDAG $ HCS.addChainsE n $ HCS.variable "x"
+   dag = Trie.elems $  BSE.dagTrie $ snd $ BSE.buildDAG $ BSE.addChains n $ BSE.variable "x"
+  in putStrLn $ show dag

### 2022-12-16 AltSharing/src/ByteStringExplicit.hs

@@ -20,11 +20,11 @@ class Exp repr where
   constant :: Int -> repr Int
 
 -- | Explicit caching of ByteStrings
-class Cacheable repr where
-  cache :: ByteString -> repr a -> repr a
+class Substitute repr where
+  subT :: ByteString -> repr a -> repr a
 
-instance Cacheable Graph where
-  cache s' (Graph g s _) = Graph g s' (Just s)
+instance Substitute Graph where
+  subT s' (Graph g s _) = Graph g s' (Just s)
 
 -- | A node is identified by a unique Int
 type NodeID = Int
@@ -39,14 +39,14 @@ data Node = NAdd NodeID NodeID
 
 -- | A Directed Acyclic Graph is inside the values of the Trie
 data DAG = DAG { dagTrie :: Trie (Node,NodeID) -- | a trie from bytestrings to Node/NodeID
-               , dagCache :: Map ByteString ByteString -- | FIXME switch to Map?
-               , dagMaxID :: Int -- | used to track # of hashcons performed
+               , dagSubMap :: Map ByteString ByteString -- | a map of labels to their subst.
+               , dagMaxID :: Int -- | used to track # of triecons performed
                } deriving Show
 
 -- | DAG construction representation via the State monad
 data Graph a = Graph { unGraph :: State DAG NodeID
                      , unStringAST :: ByteString
-                     , unAddCache :: Maybe ByteString }
+                     , unSubT :: Maybe ByteString }
 
 -- | Build a AST encoded as a ByteString
 buildStringAST :: Node -> [ByteString] -> ByteString
@@ -60,9 +60,8 @@ buildStringAST node args =
   in opString <> argsString
 
 
--- TODO call me something different then hashcons?
-hashcons :: ByteString -> Node -> State DAG NodeID
-hashcons sAST node = do
+triecons :: ByteString -> Node -> State DAG NodeID
+triecons sAST node = do
  DAG trie dCache maxID <- get
  case Trie.lookup sAST trie of
    Nothing -> let maxID' = maxID+1
@@ -74,40 +73,43 @@ hashcons sAST node = do
 seqArgs :: [Graph a] -> State DAG [NodeID]
 seqArgs inps =
   let
-    seqArg (Graph sT sAST mCache) =
-      do DAG trie cacheMap _ <- get
-         runCache sAST mCache cacheMap
-         case Trie.lookup sAST trie of
-           Nothing -> sT
[…]

### 2022-12-16 paper/paper.tex
added better plot

@@
   \includegraphics[width=0.6\textwidth]{figs/hashconscmp.png}
-  \caption{Number of calls to \mintinline{haskell}{hashcons} plotted against the number of \mintinline{haskell}{add} operations performed.} \label{fig:hashcons}
+  \caption{Number of calls to \mintinline{haskell}{hashcons} plotted against the
+    number of \mintinline{haskell}{add} operations performed.
+    Hashcons is performed without explicit sharing and is clearly exponential,
+    Triecons (without explicit sharing) and HashCons Explicit (with explicit
+    sharing) overlap and are both linear
+  } \label{fig:hashcons}
 \end{figure}
@@
 The byte string AST being built will itself suffer from lack of sharing. We're
-essentially trading extra computation for extra memory. This is often a good
-tradeoff, since memory is so plentiful in modern hardware. But under the right
-conditions it can become an issue
-
-TODO include heap profiling analysis
+essentially trading extra computation for extra memory. In our same
+\mintinline{haskell}{addChains} example from Section \ref{limithashcons}, our
+method now has exponential scaling in memory instead of computation. This can be
+a good tradeoff, since memory is so plentiful in modern hardware, but still
+presents an issue.
 
@@
 
+\section{BenchMarking}
+
 \section{Conclusion}

### 2022-12-16 AltSharing/src/ByteStringSharing.hs

@@ -104,8 +104,11 @@ plotAddChains size =
   let
     chainsDataH = map (\(x,y) -> (fromIntegral x,fromIntegral y))
       [ (n,HCS.dagNumCons $ snd $ HCS.buildDAG $ HCS.addChains n $ HCS.variable "x") | n <- [0..size] ]
+    chainsDataHS = map (\(x,y) -> (fromIntegral x,fromIntegral y))
+      [ (n,HCS.dagNumCons $ snd $ HCS.buildDAG $ HCS.addChainsE n $ HCS.variable "x") | n <- [0..size] ]
     chainsDataT = map (\(x,y) -> (fromIntegral x,fromIntegral y))
       [ (n,dagNumCons $ snd $ buildDAG $ addChains n $ variable "x") | n <- [0..size] ]
   in plot (PNG "plot.png") [Data2D [Title "Triecons",Style Linespoints,Color Blue] [] chainsDataT
+                           ,Data2D [Title "Hashcons Explicit",Style Linespoints,Color Orange] [] chainsDataHS
                            ,Data2D [Title "Hashcons",Style Linespoints,Color Red] [] chainsDataH
                            ]

### 2022-12-16 paper/paper.tex
added abstract

@@
 \begin{abstract}
-TODO The abstract should briefly summarize the contents of the paper in
-150--250 words.
+  Detection of sharing is known challenge for implementers of embedded domain
+  specific languages (DSLs). There are many solutions, each with their
+  advantages and drawbacks. Many solutions are based on observable sharing, that
+  requires either a monadic interface or use of unsafe referencing, e.g.,
+  Data.Reify. Monadic interfaces are consider unsuitable for domain experts, and
+  the use of unsafe referencing leads to fragile software.
+
+  Kiselyov's methods for implicit and explicit sharing detection for finally
+  tagless style DSLs is an elegant solution without having to resort unsafe
+  observable sharing. However these methods are not applicable to all types of
+  DSLs (including those generating hypergraphs). We will present alternative
+  methods which handle these cases. The main difference comes from the use of a
+  trie to perform hashconsing. Our method for implicit sharing essentially
+  trades worst-case exponential growth in computation for increased memory
+  footprint. To mitigate this issue, our method for explicit sharing reduces the
+  memory footprint.
 
@@
 % LocalWords:  DSLs unsafePerformIO Haskell's AST typeclass tagless Trie
-% LocalWords:  Kiselyov's haskell consing
+% LocalWords:  Kiselyov's haskell consing hypergraphs
 % Local Variables:

### 2022-12-16 paper/paper.tex
updated acknowledgements
moved christopher to the back

@@
 \author{Curtis D'Alves \and
-Christopher Kumar Anand \and
 Lucas Dutton \and
-Steven Gonder
+Steven Gonder \and
+Christopher Kumar Anand
 }
@@
 
-\subsubsection{Acknowledgements} We thank NSERC and IBM Canada for supporting this work.
+\subsubsection{Acknowledgements} We thank NSERC and IBM Canada Advanced Studies for supporting this work.
 %

### 2022-12-16 paper/paper.tex
Proof read - Lucas

@@
 and trying to adapt the let construct above,
-just creates another function function with the same problem (multiple outputs).
+just creates another function with the same problem (multiple outputs).
 
@@
 evaluated. But how do we do that without first evaluating it to gain access to
-its unique identifier. We need some way to uniquely identify it outside the monad.
+its unique identifier? We need some way to uniquely identify it outside the monad.
 
@@
 DAG.
-The byte string stays outside monad, while the DAG remains inside.
+The byte string stays outside the monad, while the DAG remains inside.
 We can do this efficiently by replacing the \mintinline{haskell}{BiMap}
@@
 \section{Explicit Sharing Of ByteString ASTs}
-We propose another solution to this issue, taking inspiration again from the
-\cite{kiselyov:sharing}, we can introduce an explicit construct for specifying
+We propose another solution to this issue, taking inspiration again from 
+\cite{kiselyov:sharing}, by introducing an explicit construct for specifying
 sharing. This time, the construct will substitute the current byte string for a
@@
 Either way it's up to the DSL
-writer to insure they don't reuse the same label.
+writer to ensure they don't reuse the same label.
 

### 2022-12-16 paper/paper.tex
typos

@@
 %
-\title{Alternative Methods for Implementing
+\title{Alternative Methods for Retaining
     Explicit and Finding Implicit Sharing in embedded DSLs}
@@
 \begin{abstract}
-  Detection of sharing is known challenge for implementers of embedded domain
+  Detection of sharing is a known challenge for implementers of embedded domain
   specific languages (DSLs). There are many solutions, each with their
@@
   requires either a monadic interface or use of unsafe referencing, e.g.,
-  Data.Reify. Monadic interfaces are consider unsuitable for domain experts, and
+  Data.Reify. Monadic interfaces are considered unsuitable for domain experts, and
   the use of unsafe referencing leads to fragile software.
@@
 this can make a substantial difference.
-However,
-one of the first things the developer will do is write a pretty printer.
+
+One of the first things the developer will do is write a pretty printer.
 That recursive function will traverse the data structure as a tree,
 and pretty print \mintinline{haskell}{exp0} twice.
-This inefficiency is a problem for code generation,
-and naive traversal of the AST does the opposite of common-subexpression elimination performed by a good optimizing compiler.
+This inefficiency is a real problem for code generation,
+and naive traversal of the AST does the opposite of the common-subexpression elimination performed by a good optimizing compiler.
 To avoid this,
@@
 The representation for the finally tagless
-instance is then a wrapper around a State monad that holds the DAG being constructed in its state and
+instance is then a wrapper around a state monad that holds the DAG being constructed in its state and
 returns the current (top) \mintinline{haskell}{NodeID}:
@@
 
-When we wrap our State monad in finally tagless style, we lose some expected sharing. In the following code, the use of the let causes the computation $x + y$ to only occur once
+When we wrap our state monad in finally tagless style, we lose some expected sharing. In the following code, the use of the let causes the computation $x + y$ to only occur once
 \begin{minted}{haskell}
@@
 Manipulating tuple values will be
-cumbersome, constantly requiring calls to custom implementations of \mintinline{haskell}{fst},\mintinline{haskell}{snd} etc. 
+cumbersome, constantly requiring calls to custom implementations of \mintinline{haskell}{fst}, \mintinline{haskell}{snd} etc. 
 And for lists you'll lose access to built-in Haskell
@@
 evaluated. But how do we do that without first evaluating it to gain access to
-its unique identifier? We need some way to uniquely identify it outside the monad.
+its unique identifier. We need some way to uniquely identify it outside the monad.
 
@@
 
-To prevent confusion, we rename the hash-consing function to \mintinline{haskell}{triecons}, as follows
+To prevent confusion, we name the hash-consing function in our method \mintinline{haskell}{triecons}:
 \begin{minted}{haskell}
@@
 We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
-fail to look up its corresponding byte string AST in the Trie.
+fail to find its corresponding serialized AST in the Trie.
 \begin{minted}{haskell}
@@
 The byte string AST being built will itself suffer from lack of sharing. We're
-essentially trading extra computation for extra memory. In our same
-\mintinline{haskell}{addChains} example from Section \ref{limithashcons}, our
+essentially trading extra computation for extra memory. In our 
+\mintinline{haskell}{addChains} example from Section~\ref{limithashcons}, our
 method now has exponential scaling in memory instead of computation. This can be
@@
 \section{Explicit Sharing Of ByteString ASTs}
-We propose another solution to this issue, taking inspiration again from 
-\cite{kiselyov:sharing}, by introducing an explicit construct for specifying
+We propose another solution to this issue, taking inspiration again from Kiselyov 
+\cite{kiselyov:sharing}, we can introduce an explicit construct for specifying
 sharing. This time, the construct will substitute the current byte string for a
@@
 different ASTs. Unfortunately, if there is a collision there's no way to escape
-the State monad to prevent or modify the substitution.
+the state monad to prevent or modify the substitution.
 In the toy example, compilation crashes, but we could catch an exception instead.
 Either way it's up to the DSL
-writer to ensure they don't reuse the same label.
+writer to insure they don't reuse the same label.
 
@@
 We have presented a method for constructing finally tagless style DSLs with
-sharing detection, that allow for multiple-output DSL functions.
+sharing detection, that allows for DSLs specifying hypergraphs (e.g., functions with multiple outputs).
 It also avoids the use of unsafe referencing as performed when doing observable sharing, c.f. \cite{gill:observablesharing}.

### 2022-12-16 paper/paper.tex

@@
 We propose another solution to this issue, taking inspiration again from Kiselyov  
-\cite{kiselyov:sharing}, we can introduce an explicit construct for specifying
+\cite{kiselyov:sharing}, by introducing an explicit construct for specifying
 sharing. This time, the construct will substitute the current byte string for a

### 2022-12-16 paper/paper.tex
added benchmarking

--- 2022-12-16 paper/paper.tex
@@
 \usepackage{minted}
+\usepackage{hyperref}
 
@@
 means DSL functions are pure, type-safe and can return Haskell container types (i.e.,
-tuples, lists, etc.) without breaking sharing.
+tuples, lists, etc.) without breaking sharing. All code will be hosted at \url{https://github.com/dalvescb/AltSharingInEDSL_Paper}
 
@@
 and trying to adapt the let construct above,
-just creates another function with the same problem (multiple outputs).
+just creates another function function with the same problem (multiple outputs).
 
@@
 DAG.
-The byte string stays outside the monad, while the DAG remains inside.
+The byte string stays outside monad, while the DAG remains inside.
 We can do this efficiently by replacing the \mintinline{haskell}{BiMap}
@@
 \section{Explicit Sharing Of ByteString ASTs}
-We propose another solution to this issue, taking inspiration again from 
-\cite{kiselyov:sharing}, by introducing an explicit construct for specifying
+We propose another solution to this issue, taking inspiration again from the
+\cite{kiselyov:sharing}, we can introduce an explicit construct for specifying
 sharing. This time, the construct will substitute the current byte string for a
@@
 
-test x y = let
+exampleSubT x y = let
   z = subT "z" (add x y)
@@
 Either way it's up to the DSL
-writer to ensure they don't reuse the same label.
+writer to insure they don't reuse the same label.
 
 \section{BenchMarking}
+
+Even with explicit sharing via substitutions, our method contains a reasonable
+amount of overhead in order to overcome the shortcomings of Kiselyov's method.
+The \mintinline{haskell}{addChains} example altered for explicit sharing with
+both methods presents a worse case scenario in terms of overhead comparison.
+Kiselyov's method is able to fully utilize it's explicit sharing and our method
+requires many substitution lookups.
+
+\begin{table}
+  \begin{tabular}{|l|l|l|l|l|l|l|l|}
+    \hline
+    {\bf Size } &  150 & 200 & 10000 & 50000 \\
+    \hline
+    {\bf HashCons time } &  0.0 secs & 0.0 secs & 0.1 secs & 0.3 secs \\
+    {\bf HashCons alloc } &  619,296 bytes & 739,304 bytes & 28,662,592 bytes & 155,993,544 bytes \\
+    {\bf TrieCons time } &   0.0 secs & 0.0 secs & 0.3 secs & 0.16 secs \\
+    {\bf TrieCons alloc } &  1,773,416 bytes & 2,333,680 bytes & 129,146,808 bytes & 723,437,504 bytes \\
+    \hline
+  \end{tabular}
+
+  \caption{Benchmarks of \mintinline{haskell}{addChains} example with full
+    explicit sharing.}\label{tab1}
+\end{table}
+
+Table \ref{tab1} gives a set of benchmarks comparing our method with Kiselyov's,
+both taking full advantage of explicit sharing. It's clear Kiselyov's method
+performs better in this situation, however it should be noted our method is
+still viable for solving very large DAG's in reasonable amounts of time / memory.
 

### 2022-12-16 AltSharing/app/Main.hs

@@ -7,7 +7,7 @@ import qualified ByteStringExplicit as BSE
 import qualified Data.Trie as Trie
 import qualified Data.Map as Map
 
-n = 2000
+n = 50000
 main :: IO ()
 main = let
    -- dag = Map.toList $ HCS.biMap $ HCS.dagMAP $ snd $ HCS.buildDAG $ HCS.addChainsE n $ HCS.variable "x"

### 2022-12-16 paper/paper.tex

--- 2022-12-16 paper/paper.tex
@@
 \title{Alternative Methods for Retaining
-    Explicit and Finding Implicit Sharing in embedded DSLs}
+    Explicit and Finding Implicit Sharing in Embedded DSLs}
 %
@@
 
+\vspace{-1cm}
 \begin{figure}
@@
 
+Even with explicit sharing via substitutions, our method contains a reasonable
+amount of overhead in order to overcome the shortcomings of Kiselyov's method.
+The \mintinline{haskell}{addChains} example altered for explicit sharing with
+both methods presents a worst case scenario in terms of overhead comparison.
+Kiselyov's method is able to fully utilize its' explicit sharing and our method
+requires many substitution lookups.
+
+\begin{table}
+  \begin{tabular}{|l|l|l|l|l|l|l|l|}
+    \hline
+    {\bf Size } &  150 & 200 & 10000 & 50000 \\
+    \hline
+    {\bf HashCons time } &  0.0 secs & 0.0 secs & 0.01 secs & 0.03 secs \\
+    {\bf HashCons alloc } &  619,296 bytes & 739,304 bytes & 28,662,592 bytes & 155,993,544 bytes \\
+    {\bf TrieCons time } &   0.0 secs & 0.0 secs & 0.03 secs & 0.16 secs \\
+    {\bf TrieCons alloc } &  1,773,416 bytes & 2,333,680 bytes & 129,146,808 bytes & 723,437,504 bytes \\
+    \hline
+  \end{tabular}
+
+  \caption{Benchmarks of \mintinline{haskell}{addChains} example with full
+    explicit sharing.}\label{tab1}
+\end{table}
+
+Table \ref{tab1} gives a set of benchmarks comparing our method with Kiselyov's,
+both taking full advantage of explicit sharing. It's clear Kiselyov's method
+performs better in this situation, however it should be noted our method is
+still viable for solving very large DAG's in reasonable amounts of time / memory.
+
 \section{Conclusion}

### 2023-02-14 paper/paper.tex
working on revisions for paper

@@
 
-Kiselyov \cite{kiselyov:sharing}  presents a method for implementing
-eDSLs in finally tagless form that generate a directed acyclic graph (DAG) with
-sharing. 
+Kiselyov \cite{kiselyov:sharing} presents a method for implementing eDSLs in
+finally tagless form that generate a directed acyclic graph (DAG) with sharing.
 However, as we will explain in sections~\ref{limithashcons} and
-\ref{limitexplicit}, for DSL functions that return multiple
-outputs (e.g., tuples, lists, etc.), Kiselyov's method of detecting sharing may require computation exponential in the size of the program,
-and his method of explicitly declaring sharing is inapplicable.
+\ref{limitexplicit}, for DSL functions that return multiple outputs (e.g.,
+tuples, lists, etc.), Kiselyov's method of implicitly detecting sharing may
+require computation exponential in the size of the program, and his method of
+explicitly declaring sharing is inapplicable.
 
@@
 \end{minted}
-the function \mintinline{haskell}{novel} exhibits this problem.
- In our work,
-this translated into the inability to process large library functions.
+the function \mintinline{haskell}{novel} will exhibit an issue in sharing of
+it's two outputs when writing it's implementation that cannot be solved using
+the methods of \cite{kiselyov:sharing}. We will illustrate this implementation
+issue in section \ref{limitexplicit}. In our work, this translated into the
+inability to process large library functions.
 
@@
 We can then create different instances to implement different functionality.
-For example, we can implement a pretty printer for our AST as
+For example, we can implement a pretty printer
 \begin{minted}{haskell}
-newtype Pretty a = Pretty { unPretty :: String }
+newtype Pretty a = Pretty { runPretty :: String }
 
@@
 
+Or generate an AST
+\begin{minted}{haskell}
+newtype AST a = AST { genAST :: Exp }
+
+instance Exp AST where
+  add x y =  AST $ Add (genAST x) (genAST y)
+  variable x = AST $ Variable x
+  constant x = AST $ Constant x
+\end{minted}
+
 Finally tagless style provides extensible, user friendly DSLs.
@@
 
-Kiselyov's method
- for detecting implicit sharing in finally tagless
-style uses hash-consing \cite{kiselyov:sharing}.
-Hash-consing is based on a bijection of nodes and a set of identifiers,
-e.g., with interface 
+The goal of detecting sharing is to generate a graph structure (like the AST in
+the previous section) that share's common subexpressions. So we're essentially
+going to generate a Directed Acyclic Graph (DAG) instead of an AST. For example
+
+\begin{minted}{haskell}
+type NodeID = Int
+data Node = NAdd NodeID NodeID
+          | NVariable String
+          | NConstant Int
+
+newtype DAG = DAG (BiMap Node) deriving Show
+\end{minted}
+
+Kiselyov's method for detecting implicit sharing in finally tagless style uses
+hash-consing \cite{kiselyov:sharing}. Hash-consing is based on a bijection of
+nodes and a set of identifiers, e.g.,
 \begin{minted}{haskell}
@@
 \end{minted}
-An efficient implementation using hashing and linear probing is given by Thai
-in his Master's thesis \cite{thai2021type}.
 
-Nodes need to be uniquely identifiable, and shouldn't be a recursive data type,
-such as
-\begin{minted}{haskell}
-type NodeID = Int
-data Node = NAdd NodeID NodeID
-          | NVariable String
-          | NConstant Int
-\end{minted}
+The method can be performed using any data structure that provides the above
+interface. An efficient implementation would use hashing and linear probing, as
+is done by Thai in his Master's thesis \cite{thai2021type}.
+
 The representation for the finally tagless
@@
 \begin{minted}{haskell}
-newtype DAG = DAG (BiMap Node) deriving Show
 
@@
 \begin{minted}{haskell}
-haskellSharing x y=
+haskellSharing x y =
  let

### 2023-02-14 paper/paper.tex
added more benchmarks to paper

@@
 the previous section) that share's common subexpressions. So we're essentially
-going to generate a Directed Acyclic Graph (DAG) instead of an AST. For example
+going to generate a Directed Acyclic Graph (DAG) instead of an AST. For example,
+we can use the following DAG structure that explicitly references nodes by a
+unique identifier
 
@@
 nodes and a set of identifiers, e.g.,
+
 \begin{minted}{haskell}
@@
 
-The representation for the finally tagless
-instance is then a wrapper around a state monad that holds the DAG being constructed in its state and
-returns the current (top) \mintinline{haskell}{NodeID}:
+In order to generate such a data structure, we will need to keep track of the
+current maximum identifier to keep them unique. The representation for the
+finally tagless instance is then a wrapper around a state monad that holds the
+DAG being constructed in its state and returns the current (top)
+\mintinline{haskell}{NodeID}:
+
 \begin{minted}{haskell}
-
 newtype Graph a = Graph { unGraph :: State DAG NodeID }
@@
 \end{minted}
+
 The trick to uncovering sharing is in the
@@
 \end{minted}
-The technique is essentially that of
-hash-consing, popularized by its use in LISP compilers, but discovered by Ershov in 1958 \cite{ershov1958:consing}. Other works have explored the use
-of type-safe hash-consing in embedded DSLs, see \cite{filliatre:typesafeconsing}.
+
+The technique is essentially that of hash-consing, popularized by its use in
+LISP compilers, but discovered by Ershov in 1958 \cite{ershov1958:consing}.
+Other works have explored the use of type-safe hash-consing in embedded DSLs,
+see \cite{filliatre:typesafeconsing}.
 
@@
 
-When we wrap our state monad in finally tagless style, we lose some expected sharing. In the following code, the use of the let causes the computation $x + y$ to only occur once
+When we wrap our state monad in finally tagless style, we lose some expected
+sharing. In the following code, the use of the let causes the computation
+$x + y$ to only occur once
 \begin{minted}{haskell}
@@
 
-Kiselyov \cite{kiselyov:sharing} recognized that the amount of computation with hash-consing ``may take a long time for large programs,'' 
-and proposed an ad-hoc solution, explicit sharing via a
-custom let construct
+Kiselyov \cite{kiselyov:sharing} recognized that the amount of computation with
+hash-consing ``may take a long time for large programs,'' and proposed an ad-hoc
+solution, explicit sharing via a custom let construct
 \begin{minted}{haskell}
@@
 
-However the method does not work for DSL functions
-returning multiple outputs via tuples or container types like lists. Recall the definition
+However the method does not work for DSL functions returning multiple outputs
+via tuples or container types like lists. If we're attempting to generate a more
+specific graph structure like a hypergraph or bi-partite graph the need to do
+this becomes more aparent. For example, in our own research we implemented a DSL
+for an instruction set architecture that generates a bipartite graph of
+instructions and the registers they act upon. Attempting to generate a graph
+where an instruction outputs multiple registers will result in returning
+multiple independent state monads.
+
+We can more concisely illustrate this problem by adding the following
+instruction to our DSL that attempts to return two separate nodes in the DAG at
+once
 \begin{minted}{haskell}
@@
 \end{minted}
-The problem is that DAG generation requires splitting the state monad in two:
+
+The issue is that DAG generation requires splitting the state monad in two:
 \begin{minted}{haskell}
@@
 \end{minted}
-Each output it returns will now have to be individually evaluated, so a
-chain of DSL functions that output 2 or more values will suffer from the same
-exponential explosion of hashcons operations,
-and trying to adapt the let construct above,
+Each output it returns will now have to be individually evaluated, so a chain of
+DSL functions that output 2 or more values will suffer from the same exponential
+explosion of hashcons operations, and trying to adapt the let construct above,
 just creates another function with the same problem (multiple outputs).
@@
 One solution to this issue is to integrate container types such as tuples and
-lists into the DSL language. However doing this eliminates the
-advantage of having an embedded language.
-Manipulating tuple values will be
-cumbersome, constantly requiring calls to custom implementations of \mintinline{haskell}{fst}, \mintinline{haskell}{snd} etc. 
-And for lists you'll lose access to built-in Haskell
-list functionality.
+lists into the DSL language. However doing this eliminates the advantage of
+having an embedded language. Manipulating tuple values will be cumbersome,
+constantly requiring calls to custom implementations of
+\mintinline{haskell}{fst}, \mintinline{haskell}{snd} etc. And for lists you'll
+lose access to built-in Haskell list functionality.
 
@@
 sharing. This time, the construct will substitute the current byte string for a
-more compact label. For safety purposes, we need to keep track of a table of
-these labels and their corresponding ASTs, to make sure we don't use the same label for different ASTs.
+more compact label.
+
 \begin{minted}{haskell}
-data DAG = DAG { dagTrie :: Trie (Node,NodeID)
-               , dagSubMap :: Map ByteString ByteString
-               , dagMaxID :: Int
-               } deriving Show
-
-data Graph a = Graph { unGraph :: State DAG NodeID
-                     , unStringAST :: ByteString
-                     , unSubT :: Maybe ByteString }
-
 class Substitute repr where
@@
 
-The \mintinline{haskell}{subT} operation substitutes the current byte string AST with a new one, and
-we define a new operation \mintinline{haskell}{subTInsert} to check if
-the label already exists in the cache map before inserting it.
+For safety purposes, we need to keep track of a table of these labels and their
+corresponding ASTs, to make sure we don't use the same label for different ASTs.
+\begin{minted}{haskell}
+data DAG = DAG { dagTrie :: Trie (Node,NodeID)
+               , dagSubMap :: Map ByteString ByteString
+               , dagMaxID :: Int
+               } deriving Show
+
+data Graph a = Graph { unGraph :: State DAG NodeID
+                     , unStringAST :: ByteString
+                     , unSubT :: Maybe ByteString }
+\end{minted}
+
+When a substitution is made via \mintinline{haskell}{subT}, the
+\mintinline{haskell}{unStringAST} field is replaced with the new label, and the
+previous serialized AST is placed in \mintinline{haskell}{unSubT}. When a
+\mintinline{haskell}{Graph} value is processed, the \mintinline{haskell}{unSubT}
+field is checked for if it contains a label
+
 \begin{minted}{haskell}
@@
   in sequence $ map seqArg inps
+\end{minted}
 
+If the \mintinline{haskell}{unSubT} field contains a label, that means the
+current \mintinline{haskell}{unStringAST} field is a substitution that needs to
+be inserted into \mintinline{haskell}{dagSubMap}. The function
+\mintinline{haskell}{subTInsert} handles this
+
+\begin{minted}{haskell}
 subTInsert :: Maybe ByteString -> ByteString
@@
 \end{minted}
+
 We need to make sure we don't attempt to insert the same substitution for two
 different ASTs. Unfortunately, if there is a collision there's no way to escape
-the state monad to prevent or modify the substitution.
-In the toy example, compilation crashes, but we could catch an exception instead.
-Either way it's up to the DSL
-writer to ensure they don't reuse the same label.
+the state monad to prevent or modify the substitution. In the toy example,
+compilation crashes, but we could catch an exception instead. Either way it's up
+to the DSL writer to ensure they don't reuse the same label as a substitution.
 
@@
 
+  \vspace{1mm}
   \caption{Benchmarks of \mintinline{haskell}{addChains} example with full
@@
 
+For more interesting benchmarks, we need to consider a more sophisticated DSL.
+We can define a DSL for an instruction set architecture like mentioned in
+section \ref{limitexplicit}. It will have the following interface
+
+\begin{minted}{haskell}
+class ISA repr where
+  -- | Load from memory into a GPR
+  ldMR :: repr MR -> Int -> (repr GPR, repr MR)
+  -- | Store a GPR into memory
+  stdMR   :: repr MR -> Int -> repr GPR -> repr MR
+  -- | Bitwise NAND of two 64-bit general-purpose registers (NNGRK)
+  nandG   :: repr GPR -> repr GPR -> repr GPR
+  -- | Bitwise NOR of two 64-bit general-purpose registers (NOGRK)
+  norG    :: repr GPR -> repr GPR -> repr GPR
+  -- | Bitwise NXOR of two 64-bit general-purpose registers (NXGRK)
+  eqvG    :: repr GPR -> repr GPR -> repr GPR
+  -- | Addition of two 64-bit general-purpose registers (AGRK)
+  addG    :: repr GPR -> repr GPR -> repr GPR
+  ...
+\end{minted}
+
+This language can be used to encode basic blocks of assembly code. These basic
+blocks may return multiple outputs, preventing us from using explicit sharing
+via let constructs. For example,
+\begin{minted}{haskell}
+add2 :: ISA repr => (repr GPR, repr GPR) -> (repr GPR, repr GPR)
+add2 (a, b) =
+    let
+        a' = addG a b
+        b' = addG a' b
+    in (a', b')
+\end{minted}
+
+We used this language to implement approximations of vector cos and tan (i.e., a
+loop body that computes over an input and output array). The loop body was
+unrolled by a factor of four making for a sufficiently large resulting DAG.
+
+We profiled cos using both Kiselyov's method (i.e., hash-consing) and are own
+(i.e., trie-consing with some explicit sharing). It wasn't possible to use
+Kiselyov's explicit sharing method where meaningful so our code was able to
+achieve greater performance, with only a small increase in memory consumption
+(see figure \ref{fig:hashvscons}).
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=0.45\textwidth]{figs/cos_profile_hashcons.png}
+  \hfill
+  \includegraphics[width=0.45\textwidth]{figs/cos_profile_triecons.png}
+  \caption{Comparison of cos using hashcons vs triecons}
+  \label{fig:hashvscons}
+\end{figure}
+
+For tan, we were unable to generate a resulting code graph using Kiselyov's
+method. Using explicit sharing via let constructs wasn't possible due to the
+multiple outputs issue and without it the amount of computation due to redundant
+hash-consing was simply too large to perform in a reasonable amount of time.
+
+We were able to generate tan using our method both without and with our explicit
+sharing, however without explicit sharing we consumed an unreasonable amount of
+memory (see figure \ref{fig:explicitvsno}). It should be noted, the main source
+of redundant computation in tan is the reuse of computed values of cos and sin.
+By simply explicitly sharing just those values we achieved the significant
+speedup shown in figure \ref{fig:explicitvsno}.
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=0.45\textwidth]{figs/noexplicit.png}
+  \hfill
+  \includegraphics[width=0.45\textwidth]{figs/explicit.png}
+  \caption{Comparison of cos using hashcons vs triecons}
+  \label{fig:explicitvsno}
+\end{figure}
+
 \section{Conclusion}
 We have presented a method for constructing finally tagless style DSLs with
-sharing detection, that allows for DSLs specifying hypergraphs (e.g., functions with multiple outputs).
-It also avoids the use of unsafe referencing as performed when doing observable sharing, c.f. \cite{gill:observablesharing}.
+sharing detection, that allows for DSLs specifying hypergraphs (e.g., functions
+with multiple outputs). It also avoids the use of unsafe referencing as
+performed when doing observable sharing, c.f. \cite{gill:observablesharing}.
 
-The method has its drawbacks in terms of memory usage, but these can be mitigated
-by explicitly specifying sharing. This does present an extra burden on the DSL
-writer to implement explicit sharing when necessary and ensure labels are not
-reused. Future work may investigate the use of a preprocessor or plugin to
-automate explicit sharing.
+The method has its drawbacks in terms of memory usage, but these can be
+mitigated by explicitly specifying sharing. This does present an extra burden on
+the DSL writer to implement explicit sharing when necessary and ensure labels
+are not reused. Future work may investigate the use of a preprocessor or plugin
+to automate explicit sharing.
 

### 2023-02-14 paper/paper.tex
updated figures

@@
   Kiselyov's methods for implicit and explicit sharing detection for finally
-  tagless style DSLs is an elegant solution without having to resort unsafe
+  tagless style DSLs is an elegant solution without having to resort to unsafe
   observable sharing. However these methods are not applicable to all types of
@@
 instance Exp Pretty where
-  add x y = Pretty $ "("++unPretty x++") + ("++unPretty y++")"
+  add x y = Pretty $ "("++runPretty x++") + ("++runPretty y++")"
   variable x = Pretty x
@@
 finally tagless instance is then a wrapper around a state monad that holds the
-DAG being constructed in its state and returns the current (top)
+DAG being constructed in it's state and returns the current (top)
 \mintinline{haskell}{NodeID}:
@@
 
-The technique is essentially that of hash-consing, popularized by its use in
+The technique is essentially that of hash-consing, popularized by it's use in
 LISP compilers, but discovered by Ershov in 1958 \cite{ershov1958:consing}.
@@
 evaluated. But how do we do that without first evaluating it to gain access to
-its unique identifier? We need some way to uniquely identify it outside the monad.
+it's unique identifier? We need some way to uniquely identify it outside the monad.
 
@@
 We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
-fail to find its corresponding serialized AST in the Trie.
+fail to find it's corresponding serialized AST in the Trie.
 \begin{minted}{haskell}
@@
   \includegraphics[width=0.45\textwidth]{figs/explicit.png}
-  \caption{Comparison of cos using hashcons vs triecons}
+  \caption{Comparison of tan using no explicit sharing vs explicit}
   \label{fig:explicitvsno}
@@
 
-The method has its drawbacks in terms of memory usage, but these can be
+The method has it's drawbacks in terms of memory usage, but these can be
 mitigated by explicitly specifying sharing. This does present an extra burden on

### 2023-02-14 paper/paper.tex
grammar fixes

@@
   methods which handle these cases. The main difference comes from the use of a
-  trie to perform hashconsing. Our method for implicit sharing essentially
+  trie to perform hash-consing. Our method for implicit sharing essentially
   trades worst-case exponential growth in computation for increased memory
@@
     number of \mintinline{haskell}{add} operations performed.
-    Hashcons is performed without explicit sharing and is clearly exponential,
-    Triecons (without explicit sharing) and HashCons Explicit (with explicit
+    Hash-consing is performed without explicit sharing and is clearly exponential,
+    Triecons (without explicit sharing) and Hashcons Explicit (with explicit
     sharing) overlap and are both linear
@@
 DSL functions that output 2 or more values will suffer from the same exponential
-explosion of hashcons operations, and trying to adapt the let construct above,
+explosion of hash-consing operations, and trying to adapt the let construct above,
 just creates another function with the same problem (multiple outputs).
@@
 This will prevent
-redundant hashconsing without the need for explicit sharing,
+redundant hash-consing without the need for explicit sharing,
 but at the expense of storing redundant byte strings.
@@
     \hline
-    {\bf HashCons time } &  0.0 secs & 0.0 secs & 0.01 secs & 0.03 secs \\
-    {\bf HashCons alloc } &  619,296 bytes & 739,304 bytes & 28,662,592 bytes & 155,993,544 bytes \\
-    {\bf TrieCons time } &   0.0 secs & 0.0 secs & 0.03 secs & 0.16 secs \\
-    {\bf TrieCons alloc } &  1,773,416 bytes & 2,333,680 bytes & 129,146,808 bytes & 723,437,504 bytes \\
+    {\bf Hash-Cons time } &  0.0 secs & 0.0 secs & 0.01 secs & 0.03 secs \\
+    {\bf Hash-Cons alloc } &  619,296 bytes & 739,304 bytes & 28,662,592 bytes & 155,993,544 bytes \\
+    {\bf Trie-Cons time } &   0.0 secs & 0.0 secs & 0.03 secs & 0.16 secs \\
+    {\bf Trie-Cons alloc } &  1,773,416 bytes & 2,333,680 bytes & 129,146,808 bytes & 723,437,504 bytes \\
     \hline
@@
 
-We used this language to implement approximations of vector cos and tan (i.e., a
+We used this language to implement approximations of vector \texttt{cos} and \texttt{tan} (i.e., a
 loop body that computes over an input and output array). The loop body was
@@
   \includegraphics[width=0.45\textwidth]{figs/cos_profile_triecons.png}
-  \caption{Comparison of cos using hashcons vs triecons}
+  \caption{Comparison of cos using hash-cons vs trie-cons}
   \label{fig:hashvscons}
@@
   \includegraphics[width=0.45\textwidth]{figs/explicit.png}
-  \caption{Comparison of tan using no explicit sharing vs explicit}
+  \caption{Comparison of tan profiling using no explicit sharing vs explicit}
   \label{fig:explicitvsno}

### 2023-02-14 paper/paper.tex
enlarged figures

@@
 the methods of \cite{kiselyov:sharing}. We will illustrate this implementation
-issue in section \ref{limitexplicit}. In our work, this translated into the
+issue in Section \ref{limitexplicit}. In our work, this translated into the
 inability to process large library functions.
@@
 
-We used this language to implement approximations of vector \texttt{cos} and \texttt{tan} (i.e., a
-loop body that computes over an input and output array). The loop body was
-unrolled by a factor of four making for a sufficiently large resulting DAG.
+We used this language to implement approximations of vector
+\mintinline{haskell}{cos} and \mintinline{haskell}{tan} (i.e., a loop body that
+computes over an input and output array). The loop body was unrolled by a factor
+of four making for a sufficiently large resulting DAG.
 
-We profiled cos using both Kiselyov's method (i.e., hash-consing) and are own
+We profiled \mintinline{haskell}{cos} using both Kiselyov's method (i.e., hash-consing) and are own
 (i.e., trie-consing with some explicit sharing). It wasn't possible to use
@@
 achieve greater performance, with only a small increase in memory consumption
-(see figure \ref{fig:hashvscons}).
+(see Fig.~\ref{fig:hashvscons}).
 
@@
   \centering
-  \includegraphics[width=0.45\textwidth]{figs/cos_profile_hashcons.png}
-  \hfill
-  \includegraphics[width=0.45\textwidth]{figs/cos_profile_triecons.png}
+  \includegraphics[width=0.6\textwidth]{figs/cos_profile_hashcons_cropped.png}
+  \includegraphics[width=0.6\textwidth]{figs/cos_profile_triecons_cropped.png}
   \caption{Comparison of cos using hash-cons vs trie-cons}
@@
 
-For tan, we were unable to generate a resulting code graph using Kiselyov's
+For \mintinline{haskell}{tan}, we were unable to generate a resulting code graph using Kiselyov's
 method. Using explicit sharing via let constructs wasn't possible due to the
@@
 
-We were able to generate tan using our method both without and with our explicit
+We were able to generate \mintinline{haskell}{tan} using our method both without and with our explicit
 sharing, however without explicit sharing we consumed an unreasonable amount of
-memory (see figure \ref{fig:explicitvsno}). It should be noted, the main source
-of redundant computation in tan is the reuse of computed values of cos and sin.
-By simply explicitly sharing just those values we achieved the significant
-speedup shown in figure \ref{fig:explicitvsno}.
+memory (see Fig.~\ref{fig:explicitvsno}). It should be noted, the main source of
+redundant computation in \mintinline{haskell}{tan} is the reuse of computed values of
+\mintinline{haskell}{cos} and \mintinline{haskell}{sin}. By simply explicitly
+sharing just those values we achieved the significant speedup shown in
+Fig.~\ref{fig:explicitvsno}.
 
@@
   \centering
-  \includegraphics[width=0.45\textwidth]{figs/noexplicit.png}
+  \includegraphics[width=0.6\textwidth]{figs/noexplicit_cropped.png}
   \hfill
-  \includegraphics[width=0.45\textwidth]{figs/explicit.png}
+  \includegraphics[width=0.6\textwidth]{figs/explicit_cropped.png}
   \caption{Comparison of tan profiling using no explicit sharing vs explicit}
@@
 
-\section{Conclusion}
+\section{Conclusion and Future Work}
 We have presented a method for constructing finally tagless style DSLs with
@@
 are not reused. Future work may investigate the use of a preprocessor or plugin
-to automate explicit sharing.
+to automate explicit sharing. Unlike an explicit let construct, it would be
+fairly straightforward to automatically bind \mintinline{haskell}{subT}
+operations to any DSL function call.
 

### 2023-02-15 paper/paper.tex
improper its
more its fixes

@@
 the function \mintinline{haskell}{novel} will exhibit an issue in sharing of
-it's two outputs when writing it's implementation that cannot be solved using
+its two outputs when writing its implementation that cannot be solved using
 the methods of \cite{kiselyov:sharing}. We will illustrate this implementation
@@
 finally tagless instance is then a wrapper around a state monad that holds the
-DAG being constructed in it's state and returns the current (top)
+DAG being constructed in its state and returns the current (top)
 \mintinline{haskell}{NodeID}:
@@
 
-The technique is essentially that of hash-consing, popularized by it's use in
+The technique is essentially that of hash-consing, popularized by its use in
 LISP compilers, but discovered by Ershov in 1958 \cite{ershov1958:consing}.
@@
 evaluated. But how do we do that without first evaluating it to gain access to
-it's unique identifier? We need some way to uniquely identify it outside the monad.
+its unique identifier? We need some way to uniquely identify it outside the monad.
 
@@
 We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
-fail to find it's corresponding serialized AST in the Trie.
+fail to find its corresponding serialized AST in the Trie.
 \begin{minted}{haskell}
@@
 both methods presents a worst case scenario in terms of overhead comparison.
-Kiselyov's method is able to fully utilize its' explicit sharing and our method
+Kiselyov's method is able to fully utilize its explicit sharing and our method
 requires many substitution lookups.
@@
 
-The method has it's drawbacks in terms of memory usage, but these can be
+The method has its drawbacks in terms of memory usage, but these can be
 mitigated by explicitly specifying sharing. This does present an extra burden on

### 2023-02-15 paper/paper.tex
updated intro

@@
 
-Kiselyov \cite{kiselyov:sharing} presents a method for implementing eDSLs in
-finally tagless form that generate a directed acyclic graph (DAG) with sharing.
+Embedded DSL's have proven useful for many applications, and there are multiple
+ways of doing the embedding. Domain experts are more comfortable with embeddings
+presented as a collection of pure functions. On the other hand, optimizing code
+generators and other downstream uses would be much easier to implement in the
+context of a monad. In particular code graph generation outside of a monad does
+not benefit from observable sharing in Haskell. Kiselyov \cite{kiselyov:sharing}
+solves this problem by presenting a method for implementing eDSLs in finally
+tagless form that generates a directed acyclic graph (DAG) with sharing.
 However, as we will explain in sections~\ref{limithashcons} and
@@
 \end{minted}
-the function \mintinline{haskell}{novel} will exhibit an issue in sharing of
-its two outputs when writing its implementation that cannot be solved using
-the methods of \cite{kiselyov:sharing}. We will illustrate this implementation
-issue in Section \ref{limitexplicit}. In our work, this translated into the
-inability to process large library functions.
+the function \mintinline{haskell}{novel} will exhibit this issue. Since it
+returns multiple outputs via a native tuple or list, it will diverge state used
+to generate the DAG that cannot be captured by Kiselyov's explicit sharing
+method. We will illustrate this implementation issue in Section
+\ref{limitexplicit}. In our work, this translated into the inability to process
+large library functions.
 

### 2023-02-15 paper/paper.tex
small fix in intro

@@
 the function \mintinline{haskell}{novel} will exhibit this issue. Since it
-returns multiple outputs via a native tuple or list, it will diverge state used
-to generate the DAG that cannot be captured by Kiselyov's explicit sharing
+returns multiple outputs via a native tuple or list, it will cause duplication
+of computation that cannot be captured by Kiselyov's explicit sharing
 method. We will illustrate this implementation issue in Section

### 2023-02-15 paper/paper.tex
small update to introduction

@@
 returns multiple outputs via a native tuple or list, it will cause duplication
-of computation that cannot be captured by Kiselyov's explicit sharing
-method. We will illustrate this implementation issue in Section
-\ref{limitexplicit}. In our work, this translated into the inability to process
-large library functions.
+of computation that cannot be captured by Kiselyov's explicit sharing method. We
+will illustrate this implementation issue in Section \ref{limitexplicit}.
+Encountering this issue in our own work with eDSLs for code generation caused a
+computational explosion on large DAGs.
 

### 2023-02-16 paper/paper.tex
last minute updates to paper

@@
 \usepackage{hyperref}
+\usepackage{float}
 
@@
 
-Finally tagless style provides extensible, user friendly DSLs.
+Finally tagless style provides extensible, user friendly DSLs. By providing one
+interface for a variety of functionality, you can choose between shallow and
+deep embeddings (see \cite{jovanovic2014yin} and
+\cite{svenningsson2013combining}) and can easily extend the DSL without having
+to alter the original class definition.
 
@@
 \mintinline{haskell}{fst}, \mintinline{haskell}{snd} etc. And for lists you'll
-lose access to built-in Haskell list functionality.
+lose access to built-in Haskell list functionality. To get the full advantage of
+embedding a DSL it should not only inherit the host language's syntactic and
+semantic structure, but also it's programming environment, as is argued in
+\cite{clements2001little}.
 
@@
 the state monad to prevent or modify the substitution. In the toy example,
-compilation crashes, but we could catch an exception instead. Either way it's up
-to the DSL writer to ensure they don't reuse the same label as a substitution.
+compilation crashes, but we could catch an exception instead if we used the
+\mintinline{haskell}{ExceptT} transformer instead of a simple state monad.
+Either way it's up to the DSL user to ensure they don't reuse the same label
+as a substitution.
 
@@
 
-We profiled \mintinline{haskell}{cos} using both Kiselyov's method (i.e., hash-consing) and are own
-(i.e., trie-consing with some explicit sharing). It wasn't possible to use
-Kiselyov's explicit sharing method where meaningful so our code was able to
-achieve greater performance, with only a small increase in memory consumption
-(see Fig.~\ref{fig:hashvscons}).
+We profiled \mintinline{haskell}{cos} using both Kiselyov's method (i.e.,
+hash-consing) and are own (i.e., trie-consing with some explicit sharing). It
+wasn't possible to use Kiselyov's explicit sharing method where meaningful so
+our code was able to achieve greater performance, 154\% speedup, with only a
+small increase in memory consumption (see Fig.~\ref{fig:hashvscons}).
 
-\begin{figure}[!h]
+\begin{figure}[H]
   \centering
@@
 
-\begin{figure}[!h]
+\begin{figure}[H]
   \centering

### 2023-02-17 paper/paper.tex
made christophers edits before submitting

@@
 
-\section{Background: Detecting Sharing}
+\section{Background: Detecting Sharing} \label{sec:background}
 
@@
 The goal of detecting sharing is to generate a graph structure (like the AST in
-the previous section) that share's common subexpressions. So we're essentially
-going to generate a Directed Acyclic Graph (DAG) instead of an AST. For example,
-we can use the following DAG structure that explicitly references nodes by a
-unique identifier
+Section \ref{sec:background}) but with sharing of common subexpressions. So
+we're going to generate a Directed Acyclic Graph (DAG) instead of an
+AST. For example, we can use the following DAG structure that explicitly
+references nodes by a unique identifier
 
@@
 However the method does not work for DSL functions returning multiple outputs
-via tuples or container types like lists. If we're attempting to generate a more
-specific graph structure like a hypergraph or bi-partite graph the need to do
-this becomes more aparent. For example, in our own research we implemented a DSL
-for an instruction set architecture that generates a bipartite graph of
-instructions and the registers they act upon. Attempting to generate a graph
-where an instruction outputs multiple registers will result in returning
-multiple independent state monads.
+via tuples or container types like lists. Generating a hypergraph or bipartite
+graph structure that allows operation nodes like \mintinline{haskell}{add} to
+have edges to multiple outputs is one way this problem will arise. For example,
+in our own research we implemented a DSL for an instruction set architecture
+that generates a bipartite graph of instructions and the registers they act
+upon. Attempting to generate a graph where an instruction outputs multiple
+values will result in returning multiple independent state monads.
 
-We can more concisely illustrate this problem by adding the following
+We can concisely illustrate this problem by adding the following
 instruction to our DSL that attempts to return two separate nodes in the DAG at
@@
 \end{minted}
-This looks a bit different because the \mintinline{haskell}{BiMap}
-was a bijective relation between nodes and node ids,
-whereas the trie maps byte strings to pairs (node,node id).
-To get the DAG expressed as a relation, project out the values of the trie.
+This looks a bit different because the \mintinline{haskell}{BiMap} was a
+bijective relation between nodes and node ids, whereas the trie maps byte
+strings to pairs (node,node id). The DAG is expressed as a relation, by
+projecting out the values of the trie.
 
@@
 For safety purposes, we need to keep track of a table of these labels and their
-corresponding ASTs, to make sure we don't use the same label for different ASTs.
+corresponding ASTs, to make sure we don't use the same label for different
+serialized ASTs.
 \begin{minted}{haskell}
@@
 \mintinline{haskell}{Graph} value is processed, the \mintinline{haskell}{unSubT}
-field is checked for if it contains a label
+field is checked to see if it contains a label
 
@@
 Even with explicit sharing via substitutions, our method contains a reasonable
-amount of overhead in order to overcome the shortcomings of Kiselyov's method.
+amount of overhead in order to overcome the limitations of Kiselyov's method.
 The \mintinline{haskell}{addChains} example altered for explicit sharing with
@@
 For more interesting benchmarks, we need to consider a more sophisticated DSL.
-We can define a DSL for an instruction set architecture like mentioned in
-section \ref{limitexplicit}. It will have the following interface
+We defined a DSL for an instruction set architecture as mentioned in
+section \ref{limitexplicit}. It will has the following interface
 
@@
 \mintinline{haskell}{cos} and \mintinline{haskell}{tan} (i.e., a loop body that
-computes over an input and output array). The loop body was unrolled by a factor
-of four making for a sufficiently large resulting DAG.
+cos or tan over an input array). The loop body was unrolled by a factor
+of four resuling in a large DAG.
 
 We profiled \mintinline{haskell}{cos} using both Kiselyov's method (i.e.,
-hash-consing) and are own (i.e., trie-consing with some explicit sharing). It
-wasn't possible to use Kiselyov's explicit sharing method where meaningful so
+hash-consing) and our own (i.e., trie-consing with some explicit sharing). It
+wasn't possible to use Kiselyov's explicit sharing method in this case so
 our code was able to achieve greater performance, 154\% speedup, with only a
@@
 multiple outputs issue and without it the amount of computation due to redundant
-hash-consing was simply too large to perform in a reasonable amount of time.
+hash-consing was simply too large, and would not terminate.
 

### 2023-04-20 paper/paper.tex
fixed typography issue after addChains example

@@
 
-\vspace{-1cm}
+% \vspace{-1cm}
 \begin{figure}
@@
 \end{figure}
+
 As shown in Fig.~\ref{fig:hashcons}, this code will perform approximately
